---
title: "Algorithm 2: Stochastic Approximation Calibration (SAC)"
author: "JoonHo Lee (jlee296@ua.edu)"
date: "`r Sys.Date()`"
description: >
  Detailed treatment of the SAC algorithm for reliability-targeted IRT
  simulation, including branch preflight, fixed-form and item-superpopulation
  estimands, Robbins-Monro updates, independent evaluation, and diagnostics.
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Algorithm 2: Stochastic Approximation Calibration (SAC)}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE, comment = "#>",
  fig.width = 7, fig.height = 5,
  fig.align = "center", out.width = "85%"
)
```

```{r setup}
library(IRTsimrel)
set.seed(42)
```

## Overview

Stochastic Approximation Calibration (SAC) is the secondary algorithm in
IRTsimrel, designed to complement EQC (Algorithm 1). SAC uses the
Robbins-Monro stochastic approximation framework to find the discrimination
scaling factor $c^*$ that achieves a target reliability.

**Reading time**: approximately 25 minutes.

**Positioning**: SAC serves both as the direct MSEM-targeting algorithm and as
a validation companion to EQC. When validating EQC, set
`reliability_metric = "info"` so both algorithms target the same estimand. SAC
also provides trajectory-based convergence diagnostics that EQC lacks.

**Prerequisites**: Familiarity with `vignette("theory-reliability")` for the
mathematical foundations. For applied workflows, see
`vignette("introduction")`.

**Key references**: Lee (2026, arXiv:2512.16012v2), Section 2.5; Robbins and
Monro (1951); Polyak and Juditsky (1992); Kushner and Yin (2003).

### When to Use SAC

| Use Case | Recommendation |
|:---------|:---------------|
| Routine simulation studies | Use EQC (faster, no tuning) |
| Independent validation of EQC | Use SAC with EQC warm start and `reliability_metric = "info"` |
| Direct MSEM targeting | Use SAC with `reliability_metric = "msem"` |
| Convergence diagnostics | SAC provides trajectory plots |
| Research on stochastic approximation | Use SAC |

The built-in Student-$t$ `heavy_tail` population is not compatible with
population MSEM targeting: its reciprocal-information expectation diverges.
SAC rejects that combination before iteration. Use the information metric or
define and label an explicitly truncated empirical design.

### Item scope is part of the target

`resample_items = FALSE` defines a **fixed-form** target. The same realized
item bank is used in preflight, every stochastic update, and independent theta
evaluation. `resample_items = TRUE` defines an **item-superpopulation** target:
fresh forms are drawn during calibration and `achieved_rho` is the mean of
form-level reliabilities over an independent evaluation bank. It is not the
reliability obtained by pooling information across forms. Result fields
`item_scope`, `estimand_signature`, and `achieved_distribution` make the
distinction explicit.

For `model = "3pl"`, lower asymptotes are supplied through
`item_params$guessing_params` and remain fixed within each item form while the
global scale multiplies discrimination only. The result stores
`guessing_vec`; `coef()` appends a `guessing` column only for 3PL results.
All-zero guessing follows the 2PL numerical path exactly. The letter $c$ below
always denotes the global discrimination scale, never the 3PL lower asymptote.


## The Robbins-Monro Framework

### The Stochastic Root-Finding Problem

After deterministic branch preflight, SAC uses stochastic approximation to
solve:

$$
g(c) = \mathbb{E}\!\left[\hat{\rho}(c)\right] - \rho^* = 0
\qquad \text{(1)}
$$

where $g(c) = \rho(c) - \rho^*$ is the reliability gap function. At each
iteration $n$, it observes a noisy evaluation:

$$
Y_n = \hat{\rho}_n(c_n) = \rho(c_n) + \varepsilon_n
\qquad \text{(2)}
$$

where $\hat{\rho}_n$ is computed from fresh theta samples at each step (and
fresh item forms only for the item-superpopulation scope),
and $\varepsilon_n$ is zero-mean noise with bounded variance:
$\mathbb{E}[\varepsilon_n \mid c_n] = 0$ and
$\mathbb{E}[\varepsilon_n^2 \mid c_n] \leq \sigma^2_\varepsilon$.

### Branch preflight

Before the first update, SAC evaluates a common-random-number pilot bank over
the configured log-scale interval. It requires a resolved, feasible, interior
increasing crossing and narrows projection bounds to that selected branch. For
item-superpopulation calibration, two disjoint form splits must select
compatible roots with overlapping branch intervals. Infeasible, decreasing,
boundary-only, non-finite, resolution-limited, or split-unstable cases stop
with an `irtsimrel_sac_*` classed error. The preflight bank is isolated from the
iteration and evaluation RNG streams.

### The Robbins-Monro Update Rule

The iterative update proceeds as:

$$
c_{n+1} = \Pi_{[c_L, c_U]}\!\left[c_n - a_n\,(\hat{\rho}_n(c_n) - \rho^*)\right]
\qquad \text{(3)}
$$

where $\Pi_{[c_L, c_U]}[\cdot]$ denotes projection onto the effective interior
of the selected branch (itself contained in the user bounds) and $a_n$ is a
decreasing step size sequence.

**Intuition**: When $\hat{\rho}_n(c_n) > \rho^*$ (current reliability is too
high), the update decreases $c_n$ (reducing discrimination, hence reliability).
When $\hat{\rho}_n(c_n) < \rho^*$, the update increases $c_n$.

### Robbins-Monro Conditions

The step size sequence must satisfy:

$$
\sum_{n=1}^{\infty} a_n = \infty
\quad\text{and}\quad
\sum_{n=1}^{\infty} a_n^2 < \infty
\qquad \text{(4)}
$$

**First condition** ($\Sigma a_n = \infty$): Ensures the algorithm can reach
any point in the parameter space, regardless of the initial value. Without this,
the total adjustment $\sum a_n \cdot |\hat{\rho}_n - \rho^*|$ might be bounded,
preventing convergence from a distant starting point.

**Second condition** ($\Sigma a_n^2 < \infty$): Controls the accumulated noise.
The total noise variance is proportional to $\sum a_n^2 \sigma^2_\varepsilon$,
which must be finite for convergence.


## Step Size Analysis

### The Step Size Sequence

SAC uses the parametric family:

$$
a_n = \frac{a}{(n + A)^\gamma}
\qquad \text{(5)}
$$

with three parameters controlling different aspects of the learning dynamics.

### Role of Each Parameter

**$a > 0$ (scale constant)**: Controls the overall magnitude of updates. Larger
$a$ means more aggressive steps, leading to faster initial progress but
potentially more oscillation.

In the package: `step_params$a` (default: 1).

**$A \geq 0$ (stabilization constant)**: Dampens early iterations where step
sizes would otherwise be very large ($a_1 = a/(1+A)^\gamma$). When $A > 0$,
the effective step size starts smaller, providing stability during the initial
transient phase. This is particularly important when `c_init` is far from $c^*$.

In the package, `step_params$A` defaults to 50. The `burn_in` argument controls
which iterations enter the Polyak-Ruppert average; it does not set `A`
automatically.

**$\gamma \in (1/2, 1]$ (decay exponent)**: Controls how quickly step sizes
decrease. In the package: `step_params$gamma` (default: 0.67).

- $\gamma = 0.51$: Slow decay, close to the boundary of the Robbins-Monro
  conditions. Maintains larger steps for longer, which can be beneficial
  when initial estimates are poor.
- $\gamma = 2/3$: A common compromise between convergence speed and stability.
- $\gamma = 1$: Fastest allowable decay. Reduces variance quickly but may
  under-correct if the initial value is far from the solution.

### Robbins-Monro Conditions Verification

For the parametric family $a_n = a/(n+A)^\gamma$:

$$
\sum_{n=1}^{\infty} \frac{a}{(n+A)^\gamma} = \infty
\quad\iff\quad \gamma \leq 1
$$

$$
\sum_{n=1}^{\infty} \frac{a^2}{(n+A)^{2\gamma}} < \infty
\quad\iff\quad 2\gamma > 1 \;\iff\; \gamma > 1/2
$$

Therefore, any $\gamma \in (1/2, 1]$ satisfies both conditions.

### Step Size Sensitivity

The effect of different step size configurations can be observed through the
convergence trajectory.

```{r step-size-sensitivity, fig.cap = "Effect of step decay parameter on SAC convergence trajectories.", fig.alt = "Multiple SAC trajectories comparing how step-size decay changes convergence behavior over iterations."}
# Common setup
eqc_ref <- eqc_calibrate(
  target_rho = 0.80, n_items = 25, model = "rasch",
  item_source = "parametric", M = 5000L, seed = 42, verbose = FALSE
)

# Run SAC with different step decay values
decays <- c(0.51, 0.67, 0.85, 1.0)
colors <- c("steelblue", "coral", "seagreen", "purple")

oldpar <- par(mar = c(4.5, 4.5, 3, 1))
on.exit(par(oldpar))

plot(NULL, xlim = c(1, 300), ylim = c(0.5, 2.5),
     xlab = "Iteration", ylab = "c_n",
     main = "SAC Trajectories: Step Decay Sensitivity")
abline(h = eqc_ref$c_star, lty = 2, col = "gray40", lwd = 1.5)

for (k in seq_along(decays)) {
  sac_k <- sac_calibrate(
    target_rho = 0.80, n_items = 25, model = "rasch",
    item_source = "parametric", reliability_metric = "info",
    c_init = 0.5,  # deliberately poor start
    n_iter = 300L, M_per_iter = 1000L,
    step_params = list(gamma = decays[k]),
    seed = 42, verbose = FALSE
  )
  lines(seq_along(sac_k$trajectory), sac_k$trajectory,
        col = colors[k], lwd = 1.5)
}

legend("topright",
       legend = c(sprintf("gamma = %.2f", decays), "EQC c*"),
       col = c(colors, "gray40"), lty = c(rep(1, 4), 2),
       lwd = c(rep(1.5, 4), 1.5), cex = 0.85)
```

**Observations**:

- Slower decay ($\gamma = 0.51$) maintains larger oscillations longer but
  eventually converges.
- Faster decay ($\gamma = 1.0$) converges quickly when starting near the
  solution, but risks under-correcting from a distant start.
- The default $\gamma = 0.67$ provides a good compromise between speed and stability.


## Polyak-Ruppert Averaging

### Why Averaging?

Under the usual locally linear stochastic-approximation conditions, a last
iterate with step size $n^{-\gamma}$ typically has mean-squared error of order
$n^{-\gamma}$. Polyak and Juditsky (1992) showed that averaging can attain the
optimal $n^{-1}$ mean-squared-error order, equivalently an
$n^{-1/2}$ error scale, under the additional regularity conditions of the CLT.

### The Averaging Formula

After discarding a burn-in period of $B$ iterations, the Polyak-Ruppert
average is:

$$
\bar{c}_N = \frac{1}{N - B} \sum_{n=B+1}^{N} c_n
\qquad \text{(6)}
$$

where $N$ is the total number of iterations and $B$ is the burn-in count.

### Convergence Rate Comparison

| Estimator | Error scale | MSE order | Notes |
|:----------|:------------|:----------|:------|
| Final iterate $c_N$ | typically $O_p(N^{-\gamma/2})$ | $O(N^{-\gamma})$ | Depends on step schedule and local conditions |
| Polyak-Ruppert $\bar{c}_N$ | $O_p(N^{-1/2})$ | $O(N^{-1})$ | Optimal under the stated averaging CLT |
| Truncated average | $O_p(N^{-1/2})$ | $O(N^{-1})$ | Requires an admissible burn-in sequence |

With $\gamma = 0.67$ (the package default), this heuristic comparison gives a
last-iterate error scale near $n^{-0.335}$ versus the averaged $n^{-0.5}$
scale. These are asymptotic orders under local assumptions, not finite-run
guarantees for every IRT reliability curve.

### Burn-In Selection

The burn-in period $B$ should be large enough to exclude the initial transient
(before the iterates have approached the solution) but small enough to
include a sufficient number of post-burn-in samples.

In the package: `burn_in` parameter, specified as an integer count (default:
`floor(n_iter / 2)`, i.e., 50% of iterations).

**Guidelines**:

- With EQC warm start: the default `burn_in = floor(n_iter / 2)` is
  sufficient (starting near the solution, transient is short).
- With cold start or APC: the default 50% burn-in is appropriate; for very
  long runs ($N > 1000$), a smaller fraction (e.g., `burn_in = floor(n_iter / 5)`)
  can be used to include more post-burn-in samples.


## Initialization Strategies

The initial value $c_0$ significantly affects SAC's convergence speed. The
package supports three initialization methods.

### Strategy 1: EQC Warm Start (Recommended)

Pass an `eqc_result` object directly to `c_init`:

An EQC object is accepted only for the same target, model, item count, metric,
and fixed-form scope. Leave `item_source` and `item_params` unspecified so SAC
can reuse the stored EQC form exactly. Supplying an EQC-info result to initialize
an MSEM run is a contract mismatch; use a numeric scale when the initializer is
only a heuristic and label the designs separately.

```{r init-eqc}
eqc_for_init <- eqc_calibrate(
  target_rho = 0.80, n_items = 25, model = "rasch",
  item_source = "parametric", M = 5000L, seed = 42, verbose = FALSE
)

fixed_rasch_init <- list(
  custom_params = list(
    beta = eqc_for_init$beta_vec,
    lambda = eqc_for_init$lambda_base
  ),
  center_difficulties = FALSE
)

sac_warm <- sac_calibrate(
  target_rho = 0.80, n_items = 25, model = "rasch",
  reliability_metric = "info", resample_items = FALSE,
  c_init = eqc_for_init,  # EQC warm start
  n_iter = 200L, M_per_iter = 1000L,
  seed = 42, verbose = FALSE,
  preflight_controls = list(M = 500L),
  evaluation_controls = list(n_forms = 3L, M = 1000L)
)

cat(sprintf("EQC c*: %.4f\n", eqc_for_init$c_star))
cat(sprintf("SAC c* (warm start): %.4f\n", sac_warm$c_star))
cat(sprintf("Init method: %s\n", sac_warm$init_method))
```

**Advantages**: Starting near the solution means the transient phase is short,
burn-in can be minimal, and fewer iterations are needed.

### Strategy 2: Analytic Pre-Calibration (APC)

When `c_init = NULL`, Rasch/2PL SAC first considers a closed-form approximation
derived under Gaussian Rasch assumptions. If it lies outside the admissible
preflight branch, the selected preflight root is used instead. For 3PL, APC is
not a valid approximation and the selected preflight root is the default
initializer.

**Definition (APC Formula).** Under the assumptions $\theta \sim N(0, 1)$ and
$\beta_i \sim N(0, \sigma^2_\beta)$ with $\lambda_i^{(0)} = 1$ (Rasch), the
approximate initial value is:

$$
c_{\text{APC}} = \sqrt{\frac{\rho^*}{I \cdot \kappa(\sigma^2) \cdot (1 - \rho^*)}}
\qquad \text{(7)}
$$

where $\sigma^2 = 1 + \sigma^2_\beta$ and the logistic-normal convolution
constant is approximated by:

$$
\kappa(\sigma^2) \approx \frac{0.25}{\sqrt{1 + \sigma^2 \pi^2 / 3}}
\qquad \text{(8)}
$$

**Derivation sketch**: Under the Gaussian Rasch model, the expected item
information at a single item is
$\mathbb{E}[\lambda^2 p(1-p)] = c^2 \kappa(\sigma^2)$. With $I$ items, the
average test information is approximately $I c^2 \kappa$, so
$\tilde{\rho} \approx \sigma^2_\theta I c^2 \kappa /
(\sigma^2_\theta I c^2 \kappa + 1)$.
Setting this equal to $\rho^*$ and solving for $c$ yields Eq. (7).

```{r init-apc}
# APC initialization (c_init = NULL triggers APC)
sac_apc <- sac_calibrate(
  target_rho = 0.80, n_items = 25, model = "rasch",
  item_source = "custom", item_params = fixed_rasch_init,
  reliability_metric = "info", resample_items = FALSE,
  c_init = NULL,  # APC warm start
  n_iter = 300L, M_per_iter = 1000L,
  seed = 42, verbose = FALSE,
  preflight_controls = list(M = 500L),
  evaluation_controls = list(n_forms = 3L, M = 1000L)
)

cat(sprintf("SAC c* (APC start): %.4f\n", sac_apc$c_star))
cat(sprintf("Init method: %s\n", sac_apc$init_method))

# Compare APC init quality
c_apc <- compute_apc_init(target_rho = 0.80, n_items = 25)
cat(sprintf("APC initial value: %.4f (vs true c* ~ %.4f)\n",
            c_apc, eqc_for_init$c_star))
```

### Strategy 3: User-Specified Value

A numeric value can be passed directly:

```{r init-user}
sac_user <- sac_calibrate(
  target_rho = 0.80, n_items = 25, model = "rasch",
  item_source = "custom", item_params = fixed_rasch_init,
  reliability_metric = "info", resample_items = FALSE,
  c_init = 1.0,  # user-specified
  n_iter = 300L, M_per_iter = 1000L,
  seed = 42, verbose = FALSE,
  preflight_controls = list(M = 500L),
  evaluation_controls = list(n_forms = 3L, M = 1000L)
)

cat(sprintf("SAC c* (user init = 1.0): %.4f\n", sac_user$c_star))
cat(sprintf("Init method: %s\n", sac_user$init_method))
```

### Initialization Comparison

```{r init-comparison, fig.cap = "Convergence speed depends strongly on initialization quality.", fig.alt = "Line plot comparing SAC scaling trajectories from EQC warm start, APC start, and user-specified starts."}
oldpar <- par(mar = c(4.5, 4.5, 3, 1))
on.exit(par(oldpar))

plot(NULL, xlim = c(1, 300), ylim = c(0, 3),
     xlab = "Iteration", ylab = "c_n",
     main = "Effect of Initialization on Convergence")
abline(h = eqc_for_init$c_star, lty = 2, col = "gray40", lwd = 1.5)

lines(seq_along(sac_warm$trajectory), sac_warm$trajectory,
      col = "steelblue", lwd = 1.5)
lines(seq_along(sac_apc$trajectory), sac_apc$trajectory,
      col = "coral", lwd = 1.5)
lines(seq_along(sac_user$trajectory), sac_user$trajectory,
      col = "seagreen", lwd = 1.5)

legend("topright",
       legend = c("EQC warm start", "APC", "User (c=1.0)", "EQC c*"),
       col = c("steelblue", "coral", "seagreen", "gray40"),
       lty = c(1, 1, 1, 2), lwd = 1.5, cex = 0.9)
```


## Convergence Theory

### Almost Sure Convergence (Theorem A.3)

**Theorem (SAC Convergence; cf. Theorem A.3, Lee 2026).** Suppose:

1. The step sizes satisfy Eq. (4): $\sum a_n = \infty$, $\sum a_n^2 < \infty$.
2. The reliability function $\rho(c)$ is Lipschitz continuous on $[c_L, c_U]$.
3. $g(c) = \rho(c) - \rho^*$ has one isolated root $c^*$ in the interior of
   the selected projected preflight branch.
4. The selected branch is globally stable over its projection interval:
   $(c-c^*)g(c)>0$ for every $c\ne c^*$ in that interval, and
   $g'(c^*)>0$ locally.
5. The noise $\varepsilon_n$ satisfies $\mathbb{E}[\varepsilon_n \mid c_n] = 0$
   and $\mathbb{E}[\varepsilon_n^2 \mid c_n] \leq \sigma^2_\varepsilon < \infty$.

Then the projected SAC iterates converge almost surely:

$$
c_n \xrightarrow{\text{a.s.}} c^* \quad \text{as } n \to \infty
\qquad \text{(9)}
$$

This is a local, branch-conditional result. It does not claim a globally
unique root on the original user interval, and it does not apply after a
reported `branch_lost` event. The configured preflight scan supplies
numerical evidence for an increasing branch; it does not prove the
branch-wide drift condition assumed by this theorem.

**Proof sketch.** The proof follows the classical Robbins-Monro theory,
extended to the projected (constrained) setting by Kushner and Yin (2003,
Chapter 5). The key steps are:

1. **Lyapunov function**: Define $V(c) = (c - c^*)^2$. Compute the
   conditional expectation of $V(c_{n+1})$ given $c_n$:
   $$
   \begin{aligned}
   \mathbb{E}[V(c_{n+1}) \mid c_n]
   &= V(c_n) - 2a_n(c_n - c^*) g(c_n) + a_n^2 \mathbb{E}[(\hat{\rho}_n - \rho^*)^2 \mid c_n]
   \end{aligned}
   $$

2. **Drift condition**: Since $g(c)(c - c^*) > 0$ for $c \neq c^*$ (the
   reliability function is increasing, so $g$ has the same sign as
   $c - c^*$), the drift term $-2a_n(c_n - c^*)g(c_n) < 0$ provides
   contraction toward $c^*$.

3. **Noise control**: The accumulated noise $\sum a_n^2 \sigma^2_\varepsilon <
   \infty$ (by the step size conditions), so the noise does not prevent
   convergence.

4. **Almost sure convergence**: By the Robbins-Siegmund theorem (a
   supermartingale convergence result), $V(c_n) \to V^*$ a.s. for some
   random variable $V^*$. The drift condition then forces $V^* = 0$, giving
   $c_n \to c^*$ a.s.

### Asymptotic Normality of Polyak-Ruppert Average

**Theorem (Polyak-Ruppert CLT).** Under the conditions of Theorem A.3 and
additional smoothness assumptions, the Polyak-Ruppert average satisfies:

$$
\sqrt{N - B}\,(\bar{c}_N - c^*)
\xrightarrow{d}
N\!\left(0,\, \frac{\sigma^2_\varepsilon}{g'(c^*)^2}\right)
\qquad \text{(10)}
$$

Under the additional regular stochastic-approximation conditions of
Polyak--Juditsky theory, this covariance is asymptotically efficient for the
local stochastic-approximation problem. It is not an unconditional
Cramer--Rao claim for arbitrary IRT designs or noise processes.

**Rate comparison**: Under the same local conditions, $\gamma=0.67$ gives a
typical last-iterate MSE order $O(n^{-0.67})$, whereas the Polyak--Ruppert
average has $O(n^{-1})$ MSE and the asymptotic covariance shown above. This is
the rate improvement; the reverse ordering reported in earlier documentation
was incorrect.

### Convergence Diagnostics

SAC provides automatic convergence assessment through several statistics.

```{r convergence-diag}
# Run SAC with enough iterations
sac_diag <- sac_calibrate(
  target_rho = 0.80, n_items = 25, model = "rasch",
  reliability_metric = "info", resample_items = FALSE,
  c_init = eqc_for_init, n_iter = 300L, M_per_iter = 1000L,
  seed = 42, verbose = FALSE,
  preflight_controls = list(M = 500L),
  evaluation_controls = list(n_forms = 3L, M = 1000L)
)

conv <- sac_diag$convergence
cat("Convergence diagnostics:\n")
cat(sprintf("  Converged:          %s\n", conv$converged))
cat(sprintf("  Status:             %s\n", sac_diag$calibration_status))
cat(sprintf("  Status flags:       %s\n", conv$status_flags))
cat(sprintf("  Mean (first half):  %.4f\n", conv$mean_first_half))
cat(sprintf("  Mean (second half): %.4f\n", conv$mean_second_half))
cat(sprintf("  SD (post-burn-in):  %.4f\n", conv$sd_post_burn))
cat(sprintf("  Hit lower bound:    %s\n", conv$hit_lower_bound))
cat(sprintf("  Hit upper bound:    %s\n", conv$hit_upper_bound))
```

**Convergence criterion**: The split-mean trajectory diagnostic is:

$$
\frac{|\bar{c}_{\text{first half}} - \bar{c}_{\text{second half}}|}
{\max(|c^*|, \epsilon)}
< 0.05
$$

This mean-split diagnostic detects systematic drift, which would indicate
that the iterates have not yet settled around the solution. The threshold is
relative to the calibrated scale because the diagnostic is measured on the
\(c\) scale; \(\epsilon\) is machine precision.

The public `convergence$converged` flag requires all of the following:

1. enough post-burn-in iterates and the split-mean diagnostic below 5%;
2. independent achieved reliability within 0.05 of the target; and
3. no `branch_lost` condition (including repeated branch-bound hits, an
   out-of-branch final average, or a nonpositive final branch slope).

Thus the split-mean statistic alone is not the convergence decision.

**Post-burn-in SD**: Should be small relative to $c^*$. Values above 0.1
suggest more iterations or better initialization.

`iteration_trace` is the authoritative update ledger. Each row pairs the
pre-update `c_evaluated` with `rho_update`, then records `c_raw`, `c_updated`,
the reliability recomputed at that updated scale (`rho_updated`), and any
projection. The top-level `trajectory` and `rho_trajectory` are the aligned
post-update scale/reliability sequence; `evaluation_trajectory` (also exposed as
`rho_scale_trajectory`) and `rho_update_trajectory` preserve the aligned
pre-update pair. This removes the one-step pairing ambiguity present in earlier
schemas.


## Comparative Analysis: EQC vs SAC

### When They Agree

When both algorithms use the same reliability metric and a verifiably
comparable design, and the SAC run has converged, their scales can be compared.
`compare_eqc_sac()` checks the estimand signature, model, item count, item
scope, fixed bank or normalized generator specification, scale convention, and
calibration status. If comparability cannot be established, it retains the
side-by-side numbers but returns `agreement = NA` and explanatory
`comparability_reasons`; a percentage threshold is not evidence across different
estimands.

```{r agree-demo}
# Both targeting rho* = 0.80 with info metric
eqc_comp <- eqc_calibrate(
  target_rho = 0.80, n_items = 25, model = "rasch",
  item_source = "parametric", reliability_metric = "info",
  M = 5000L, seed = 42, verbose = FALSE
)

sac_comp <- sac_calibrate(
  target_rho = 0.80, n_items = 25, model = "rasch",
  reliability_metric = "info", resample_items = FALSE,
  c_init = eqc_comp, n_iter = 300L, M_per_iter = 1000L,
  seed = 42, verbose = FALSE,
  preflight_controls = list(M = 500L),
  evaluation_controls = list(n_forms = 3L, M = 1000L)
)

comparison <- compare_eqc_sac(eqc_comp, sac_comp, verbose = FALSE)
cat(sprintf("EQC c*:   %.4f\n", comparison$c_eqc))
cat(sprintf("SAC c*:   %.4f\n", comparison$c_sac))
cat(sprintf("Diff:     %.4f (%.2f%%)\n", comparison$diff_abs, comparison$diff_pct))
cat(sprintf("Agree:    %s\n", comparison$agreement))
```

### Sources of Disagreement

When EQC and SAC disagree by more than 5%, investigate the following:

1. **Different reliability metrics**: For EQC/SAC agreement checks, ensure
   SAC uses `reliability_metric = "info"`. Direct SAC `"msem"` targeting is a
   different estimand, so disagreement with EQC-info is expected.

2. **SAC non-convergence**: Check `sac_result$convergence$converged`. If
   `FALSE`, increase `n_iter` or use EQC warm start.

3. **Small M in EQC**: Low quadrature size introduces Monte Carlo error.
   Increase `M` in EQC.

4. **Different item scope/design**: Fixed-form EQC and a resampled-form SAC
   target different item measures. Inspect `comparable` and
   `comparability_reasons` rather than interpreting their raw percentage gap.

5. **Different root branch**: Confirm the selected root direction and policy.
   A high-scale decreasing root is not interchangeable with the default lowest
   increasing root.

6. **Extreme target**: Very high ($> 0.95$) or very low ($< 0.50$) targets
   have flatter reliability curves, making both algorithms less precise.

### What to Do When They Disagree

```{r disagree-strategy, eval = FALSE}
# Strategy 1: Increase SAC iterations
sac_longer <- sac_calibrate(
  ..., n_iter = 500L, seed = 42
)

# Strategy 2: Increase EQC quadrature
eqc_precise <- eqc_calibrate(
  ..., M = 50000L, seed = 42
)

# Strategy 3: Verify same metric
stopifnot(eqc_result$metric == sac_result$metric)

# Strategy 4: Run multiple SAC replications
sac_reps <- replicate(5, {
  sac_calibrate(
    ..., seed = sample.int(10000, 1), verbose = FALSE
  )$c_star
})
cat(sprintf("SAC c* across 5 reps: mean = %.4f, sd = %.4f\n",
            mean(sac_reps), sd(sac_reps)))
```


## Diagnostic Examples

### Trajectory Analysis

The `plot()` method provides visual diagnostics for SAC convergence.

```{r trajectory-plot, fig.cap = "SAC convergence trajectory showing the scaling factor iterations.", fig.alt = "SAC diagnostic plot showing scaling factor values across stochastic approximation iterations."}
# Run SAC from APC start for a more interesting trajectory
sac_traj <- sac_calibrate(
  target_rho = 0.80, n_items = 25, model = "rasch",
  item_source = "parametric", reliability_metric = "info",
  c_init = NULL,  # APC start
  n_iter = 300L, M_per_iter = 1000L,
  seed = 42, verbose = FALSE
)

plot(sac_traj, type = "c")
```

```{r rho-trajectory-plot, fig.cap = "SAC reliability trajectory showing convergence to the target.", fig.alt = "SAC diagnostic plot showing reliability estimates fluctuating around the target over iterations."}
plot(sac_traj, type = "rho")
```

### Replication Variability

SAC is inherently stochastic. Running multiple replications reveals the
variability of the estimate.

```{r replication-var}
n_reps <- 8
sac_reps <- numeric(n_reps)

for (r in seq_len(n_reps)) {
  sac_r <- sac_calibrate(
    target_rho = 0.80, n_items = 25, model = "rasch",
    reliability_metric = "info", resample_items = FALSE,
    c_init = eqc_for_init,
    n_iter = 200L, M_per_iter = 1000L,
    seed = r * 100, verbose = FALSE,
    preflight_controls = list(M = 500L),
    evaluation_controls = list(n_forms = 3L, M = 1000L)
  )
  sac_reps[r] <- sac_r$c_star
}

cat("SAC replication study (8 runs with EQC warm start):\n")
cat(sprintf("  Mean c*:  %.4f\n", mean(sac_reps)))
cat(sprintf("  SD c*:    %.4f\n", sd(sac_reps)))
cat(sprintf("  Range:    [%.4f, %.4f]\n", min(sac_reps), max(sac_reps)))
cat(sprintf("  EQC c*:   %.4f\n", eqc_for_init$c_star))
cat(sprintf("  |Mean - EQC|: %.4f\n", abs(mean(sac_reps) - eqc_for_init$c_star)))
```

### Comparison Across Different Conditions

```{r conditions-compare}
# Test across different target reliabilities
targets <- c(0.60, 0.70, 0.80, 0.90)
cat("EQC vs SAC across target reliabilities (25 Rasch items):\n")
cat(sprintf("  %-8s %-10s %-10s %-10s %-8s\n",
            "Target", "EQC c*", "SAC c*", "|Diff|", "Agree"))

for (rho_t in targets) {
  eqc_t <- eqc_calibrate(
    target_rho = rho_t, n_items = 25, model = "rasch",
    item_source = "parametric", reliability_metric = "info",
    M = 5000L, seed = 42, verbose = FALSE
  )
  sac_t <- sac_calibrate(
    target_rho = rho_t, n_items = 25, model = "rasch",
    reliability_metric = "info", resample_items = FALSE,
    c_init = eqc_t, n_iter = 200L, M_per_iter = 1000L,
    seed = 42, verbose = FALSE,
    preflight_controls = list(M = 500L),
    evaluation_controls = list(n_forms = 3L, M = 1000L)
  )
  diff_abs <- abs(eqc_t$c_star - sac_t$c_star)
  diff_pct <- 100 * diff_abs / eqc_t$c_star
  agree <- diff_pct < 5
  cat(sprintf("  %-8.2f %-10.4f %-10.4f %-10.4f %-8s\n",
              rho_t, eqc_t$c_star, sac_t$c_star, diff_abs,
              ifelse(agree, "YES", "NO")))
}
```

### Effect of M_per_iter

The per-iteration Monte Carlo sample size controls the noise level in each
SAC step.

```{r m-per-iter}
m_values <- c(200, 500, 1000, 2000)
cat("SAC sensitivity to M_per_iter (target = 0.80, 300 iter):\n")
cat(sprintf("  %-12s %-10s %-12s %-10s\n",
            "M_per_iter", "c*", "Post-burn SD", "Converged"))

for (m_val in m_values) {
  sac_m <- sac_calibrate(
    target_rho = 0.80, n_items = 25, model = "rasch",
    reliability_metric = "info", resample_items = FALSE,
    c_init = eqc_for_init, n_iter = 300L,
    M_per_iter = as.integer(m_val),
    seed = 42, verbose = FALSE,
    preflight_controls = list(M = 500L),
    evaluation_controls = list(n_forms = 3L, M = 1000L)
  )
  cat(sprintf("  %-12d %-10.4f %-12.4f %-10s\n",
              m_val, sac_m$c_star,
              sac_m$convergence$sd_post_burn,
              sac_m$convergence$converged))
}
```

### Independent achieved-reliability evaluation

Calibration and final evaluation use separate deterministic RNG streams. For a
fixed form, independent theta blocks quantify integration variability and the
concatenated evaluation nodes define `achieved_rho`. For an
item-superpopulation target, each evaluation block has an independently drawn
form and `achieved_rho` is the arithmetic mean of the form reliabilities;
`achieved_se` and `achieved_distribution` summarize their dispersion.
`representative_achieved_rho` and the stored representative item bank remain
available for inspection, but they are not substitutes for the canonical
superpopulation mean. Likewise, `predict(object)` returns the stored aggregate,
whereas prediction at explicit new scales is labeled with its prediction scope.

The result also records `preflight`, `branch`, `iteration_trace`,
`calibration_design`, `evaluation_design`, and `rng_provenance`. These fields
should accompany a reproducibility report; `c_star` alone does not identify the
estimand or branch.


## Deprecated Alias

The function was previously named `spc_calibrate()`. For backward
compatibility, `spc_calibrate()` still works but is deprecated and will
emit a warning. Use `sac_calibrate()` in all new code.

Similarly, `compare_eqc_spc()` is deprecated in favor of
`compare_eqc_sac()`.


## Summary

SAC is a theoretically grounded validation companion to EQC:

| Aspect | Summary |
|:-------|:--------|
| **Method** | Robbins-Monro stochastic approximation |
| **Convergence** | Almost sure (Theorem A.3) |
| **Optimal rate** | $O(n^{-1/2})$ via Polyak-Ruppert averaging |
| **Key advantage** | Direct MSEM targeting and independent validation of EQC |
| **Key disadvantage** | Requires tuning (step size, iterations, burn-in) |
| **Default metric** | MSEM-based (`"msem"` / `"bar"`); use `"info"` for EQC validation |
| **Item target** | Fixed form or mean of form reliabilities, selected explicitly |
| **Best practice** | Inspect preflight/branch and independent achieved distribution; use a comparable EQC warm start only for fixed-form info calibration |
| **Convergence check** | `sac_result$convergence$converged` |


## References

Lee, J.-H. (2026). Reliability-Targeted Simulation of Item Response Data:
Solving the Inverse Design Problem. arXiv:2512.16012v2.
https://doi.org/10.48550/arXiv.2512.16012

Robbins, H., & Monro, S. (1951). A stochastic approximation method. *The Annals
of Mathematical Statistics, 22*(3), 400--407.

Polyak, B. T., & Juditsky, A. B. (1992). Acceleration of stochastic
approximation by averaging. *SIAM Journal on Control and Optimization, 30*(4),
838--855.

Kushner, H. J., & Yin, G. G. (2003). *Stochastic Approximation and Recursive
Algorithms and Applications* (2nd ed.). Springer.

Spall, J. C. (2003). *Introduction to Stochastic Search and Optimization*.
Wiley.
