---
title: "Algorithm 1: Empirical Quadrature Calibration (EQC)"
author: "JoonHo Lee (jlee296@ua.edu)"
date: "`r Sys.Date()`"
description: >
  Detailed treatment of the EQC algorithm for reliability-targeted IRT
  simulation, including adaptive root-topology discovery, branch selection,
  Monte Carlo approximation, numerical stability, and diagnostics.
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Algorithm 1: Empirical Quadrature Calibration (EQC)}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE, comment = "#>",
  fig.width = 7, fig.height = 5,
  fig.align = "center", out.width = "85%"
)
```

```{r setup}
library(IRTsimrel)
set.seed(42)
```

## Overview

Empirical Quadrature Calibration (EQC) is the primary algorithm in IRTsimrel
for solving the inverse reliability problem. Given a target marginal
reliability $\rho^*$, EQC finds a global discrimination scaling factor $c^*$
such that a fixed-form empirical approximation to population
average-information reliability equals the target.

**Reading time**: approximately 25 minutes.

**Prerequisites**: This vignette assumes familiarity with the mathematical
foundations developed in `vignette("theory-reliability")`. For applied usage
tutorials and workflows, see `vignette("introduction")`.

**Key references**: Lee (2026, arXiv:2512.16012v2), Sections 2.3--2.4;
Brent (1973).

### Why EQC is the Recommended Default

| Property | Benefit |
|:---------|:--------|
| Deterministic | Same inputs always produce same output (given seed) |
| Explicit topology | Detects multiple roots and local extrema before selection |
| No tuning parameters | No step sizes, learning rates, or burn-in to configure |
| Verified bracketing | Ordinary sign-changing crossings are locally polished when the bracket solver succeeds |
| Reproducible | Fixed theta nodes and item form for a given seed |


## Algorithm Statement

EQC solves the scalar root-finding problem $\hat{\rho}_M(c) = \rho^*$ using
a three-step procedure.

### Formal Algorithm

**Algorithm 1** (Empirical Quadrature Calibration).

> **Input**: Target reliability $\rho^*$, number of items $I$, latent
>   distribution $G$, item parameter distribution $H$, quadrature size $M$,
>   search bounds $[c_L, c_U]$, tolerance $\varepsilon$.
>
> **Step 1** (Quadrature Sampling).
>   Draw $\{\theta_m\}_{m=1}^M \sim G$ and
>   $\{(\beta_i, \lambda_i^{(0)})\}_{i=1}^I \sim H$.
>   Fix these samples for all subsequent evaluations.
>
> **Step 2** (Empirical Reliability Function).
>   Define $\hat{\rho}_M : (0, \infty) \to (0, 1)$ by:
>
>   $$
>   \hat{\rho}_M(c) = \rho\bigl(c;\, \{\theta_m\}_{m=1}^M,\,
>   \boldsymbol{\beta},\, c \cdot \boldsymbol{\lambda}^{(0)}\bigr)
>   $$
>
>   using the bracket-checked average-information reliability
>   $\tilde{\rho}$ (`reliability_metric = "info"` / `"tilde"`).
>
> **Step 3** (Topology, Polishing, and Selection).
>   Scan $x=\log(c)$ adaptively on the exact interval $[c_L,c_U]$, refine
>   detected extrema and target features, polish every sign-changing bracket
>   with Brent's method, and select a root with `root_policy`. The default
>   selects the lowest increasing crossing.
>
> **Output**: Calibrated scaling factor $\hat{c}^*_M$, achieved reliability
>   $\hat{\rho}_M(\hat{c}^*_M)$, calibrated item parameters.

### Why Fixed Quadrature?

A critical design choice is that the same Monte Carlo samples
$\{\theta_m\}$ and $\{(\beta_i, \lambda_i^{(0)})\}$ are used for every
evaluation of $\hat{\rho}_M(c)$ during the root-finding iteration. This makes
$\hat{\rho}_M(c)$ a _deterministic_ function of $c$ for a given draw, which:

1. Makes every scanner and root-polishing evaluation reproducible.
2. Eliminates the need for step size tuning.
3. Produces reproducible results for a given seed.

The trade-off is that the solution $\hat{c}^*_M$ depends on the particular
quadrature draw, introducing Monte Carlo error of order $O(1/\sqrt{M})$.


## Adaptive Topology Scan and Brent Polishing

EQC does not call one global root solver and assume monotonicity. It first
constructs an adaptive log-scale map of the configured interval. Boundaries,
detected extrema, crossings, tangencies, and plateaus are stored in
`result$misc$topology`. Brent's method (Brent, 1973) is then attempted
locally for each sign-changing bracket. Successfully polished ordinary
crossings retain their solver diagnostic; a near-target grid fallback is
recorded separately when polishing fails.

The interval is never expanded automatically. Topology is described as
"detected at the configured resolution" because no finite grid can guarantee
detection of an arbitrarily narrow feature between all sampled points.

### Components of Brent's Method

Brent's method adaptively selects among three strategies at each iteration:

1. **Bisection**: Takes the midpoint of the current bracketing interval
   $[a, b]$. Always converges but is slow (linear rate).

2. **Secant method**: Uses linear interpolation between the two most recent
   points. Superlinear convergence rate of approximately $\varphi \approx 1.618$
   (the golden ratio), but not guaranteed to stay within bounds.

3. **Inverse quadratic interpolation (IQI)**: Fits a quadratic through the
   three most recent points (in the inverse direction). Under the usual
   smoothness and simple-root conditions, pure IQI has asymptotic order
   $p \approx 1.8393$, the positive root of
   $p^3 = p^2 + p + 1$.

At each step, Brent's method:

- Attempts IQI or secant first (for speed).
- Falls back to bisection if the superlinear step would leave the bracket
  or make insufficient progress.

### Convergence Properties

**Proposition (Bracket contraction).** Given a continuous function $g$ on
$[a,b]$ with $g(a)g(b)<0$, the bisection safeguard can reduce the
root-containing bracket to a requested scale-width tolerance in:

$$
O\!\left(\log_2\!\left(\frac{b-a}{\varepsilon_c}\right)\right)
\text{ iterations (worst-case bracket contraction)}
$$

Here $\varepsilon_c$ is a tolerance on root location or bracket width, not an
automatic bound on $|g(c)|$. Continuity implies the residual tends to zero as
the bracket contracts around a root, but a finite residual guarantee requires
additional information such as a modulus of continuity or a local Lipschitz
bound.

In practice, accepted interpolation steps often converge much faster than
pure bisection. For pure IQI, the local error follows a three-step product
recurrence under regularity conditions, yielding the superlinear order
$p \approx 1.8393$ above. Brent's hybrid retains the bisection safeguard; it
does not have quadratic or doubly exponential convergence as a general
guarantee.

### Why Brent's Method is Used Locally

Several properties of the reliability function make Brent's method
particularly well-suited:

1. **Bracketed root**: Every ordinary crossing passed to `uniroot()` already
   has a verified local sign change. Several brackets may coexist.

2. **Smooth function**: $\hat{\rho}_M(c)$ is infinitely differentiable in $c$,
   enabling fast superlinear convergence.

3. **No derivative needed**: Unlike Newton's method, Brent's method does not
   require computing $\partial \hat{\rho}_M / \partial c$, avoiding the
   complexity of differentiating through the Monte Carlo sum.

4. **Robustness**: The bisection fallback prevents divergence even when the
   function has near-zero slope (high or low reliability regions).

### Root policies and statuses

`root_policy = "lowest_increasing"` is the scientific default: among detected
positive-slope crossings it chooses the smallest scale, avoiding an implicit
switch to a high-scale saturation branch. `"nearest_increasing"` uses $c=1$
as its reference. `"lowest_any"` and `"highest_any"` are expert policies that
can select decreasing crossings; tangencies and plateaus require separate
explicit opt-in.

Feasibility, topology quality, root kind/direction, and calibration outcome
are stored as separate fields. A boundary hit is not mislabeled as an interior
root. If the target is outside the detected reliability range, EQC returns the
best-achievable scale with an infeasibility warning. If a feasible target has
no root admissible under the policy, or the topology is non-finite/unresolved,
EQC raises a classed error.


## Convergence Theory

The theoretical properties of EQC are established in Lee (2026), Appendix A.
Here we state the main results with proof sketches. For the full notation and
definitions, see `vignette("theory-reliability")`.

### Consistency (Theorem A.1)

**Theorem (Branch-conditional consistency of EQC).** Let $c^*$ be an isolated
population solution $\tilde{\rho}(c^*) = \rho^*$ on a specified increasing
branch, and let $\hat{c}^*_M$ be the matching EQC root with quadrature size
$M$. Then, under the regularity and root-separation conditions below:

$$
\hat{c}^*_M \xrightarrow{\text{a.s.}} c^* \quad \text{as } M \to \infty
$$

**Proof sketch.** The argument proceeds in three steps:

1. **Uniform convergence**: By the Uniform Law of Large Numbers (ULLN), the
   empirical reliability function converges uniformly over the compact set
   $[c_L, c_U]$:
   $$
   \sup_{c \in [c_L, c_U]} |\hat{\rho}_M(c) - \rho(c)| \xrightarrow{\text{a.s.}} 0
   $$
   This requires verifying that the family
   $\{h_c(\theta, \boldsymbol{\beta}, \boldsymbol{\lambda})\}_{c \in [c_L, c_U]}$
   satisfies a Lipschitz or bounded variation condition in $c$, which follows
   from the smoothness of the logistic function.

2. **Local uniqueness of the root**: On the population target branch described in
   Proposition 1 in `vignette("theory-reliability")`, the equation
   $\rho(c) = \rho^*$ has one isolated solution $c^*$. Other branches may
   contain other roots. The finite empirical implementation enumerates local
   brackets and applies the stated selection rule.

3. **Root convergence**: Combining uniform convergence of the objective with
   uniqueness of the root, the Argmax Continuous Mapping Theorem (or its
   zero-finding analogue) yields $\hat{c}^*_M \to c^*$ a.s.

### Asymptotic Normality (Theorem A.2)

**Theorem (Asymptotic Normality of EQC).** Under regularity conditions:

$$
\sqrt{M}\,(\hat{c}^*_M - c^*) \xrightarrow{d} N\!\left(0,\, \frac{\sigma^2_g}{g'(c^*)^2}\right)
$$

where $g(c) = \rho(c) - \rho^*$, $g'(c^*) = \rho'(c^*)$ is the slope of the
reliability function at the solution, and $\sigma^2_g$ is the asymptotic
variance of $\hat{g}_M(c^*)$.

**Proof sketch.** Apply the delta method (implicit function theorem version).
By the CLT:

$$
\sqrt{M}\,\hat{g}_M(c^*) \xrightarrow{d} N(0, \sigma^2_g)
$$

Since $\hat{c}^*_M$ solves $\hat{g}_M(\hat{c}^*_M) = 0$, a Taylor expansion
around $c^*$ gives:

$$
0 = \hat{g}_M(\hat{c}^*_M) \approx \hat{g}_M(c^*) + g'(c^*)\,(\hat{c}^*_M - c^*)
$$

Solving for $\hat{c}^*_M - c^*$ and scaling by $\sqrt{M}$ yields the result.
The key requirement is $g'(c^*) \neq 0$, supplied by strict local
monotonicity on the selected branch.

### Monte Carlo Error Analysis

**Practical implications.** The asymptotic normality result (above) implies
that the Monte Carlo standard error of $\hat{c}^*_M$ is:

$$
\text{SE}(\hat{c}^*_M) \approx \frac{\sigma_g}{\sqrt{M}\, |g'(c^*)|}
$$

The reliability function's slope $|g'(c^*)|$ acts as an amplification factor:
steeper slopes (smaller $c^*$, moderate reliability targets) produce smaller
estimation errors, while flat slopes (extreme reliability targets near 0 or 1)
produce larger errors.

For the reliability _estimate_ itself, by the delta method:

$$
|\hat{\rho}_M(\hat{c}^*_M) - \rho^*| = O_p\!\left(\frac{1}{\sqrt{M}}\right)
$$


## Numerical Stability

### Stable log-information reductions

The probability and information kernels are evaluated with stable logistic
identities, including extreme linear predictors. Test information is reduced
on the log scale and MSEM uses log-sum-exp arithmetic. The implementation does
not impose a `1e-10` information floor: clipping would turn a divergent or
under-resolved reciprocal-information estimand into a plausible-looking finite
number. Non-finite diagnostics are surfaced instead.

### Search Bound Sensitivity

The choice of $[c_L, c_U]$ affects both feasibility and numerical behavior:

- **Too narrow**: The target $\rho^*$ may fall outside the detected achievable
  range on $[c_L, c_U]`; EQC returns the best-achievable point with an explicit
  status and warning.
- **Too wide**: Finite-grid saturation can create additional extrema and
  crossings even for the information metric. The scanner enumerates them, but
  a wider interval costs more evaluations and makes the root policy important.

**Default recommendation**: `c_bounds = c(0.3, 3)` works well for most
applications. Extend to `c(0.1, 5)` or `c(0.1, 10)` for high-reliability
targets ($\rho^* > 0.90$) or unusual item/latent configurations.

### Edge Cases

| Scenario | Behavior | Recommendation |
|:---------|:---------|:---------------|
| Target outside detected range | Returns best-achievable $c$ with warning/status | Inspect `misc$topology`; revise design or bounds |
| Several admissible roots | Selects by `root_policy` and records all roots | Report policy and selected direction |
| Tangency or plateau only | No default admissible root | Opt in only with a scientific justification |
| Scan budget/non-finite gap | Classed topology error | Inspect controls and estimand integrability |
| `n_items = 1` | Very flat reliability curve | Use larger `M` for precision |


## Reliability Metric Comparison

EQC targets the average-information reliability metric. The MSEM-based metric
is still important for interpretation and cross-evaluation, but direct
`"msem"` / `"bar"` targeting is handled by SAC because the MSEM objective can
be non-monotone under scalar root finding.

### Average-Information (rho-tilde): `reliability_metric = "info"` {#average-information-tilderho-reliability_metric-info}

**Advantages**:

- Stable on practical EQC search intervals; on wider finite empirical bounds,
  EQC scans and refines all detected extrema before declaring infeasibility.
- Faster convergence of Brent's method (steeper slope in typical range).
- Recommended by Lee (2026) as the default.

**Interpretation**: The coefficient obtained by applying the reliability
transform to average test information. On the same finite information grid,
Jensen's inequality makes it an upper bound on the MSEM-based coefficient.
The two are distinct estimands; this ordering is not a bias claim that one is
the universal `"true"` reliability.

### MSEM-Based (w-bar): cross-evaluation, not EQC targeting {#msem-based-barw-cross-evaluation-not-eqc-targeting}

**Advantages**:

- MSEM-based reliability using the reciprocal-information approximation to
  conditional error variance.
- Accounts for heterogeneous measurement precision across $\theta$.

**Caution**: Can be non-monotone at extreme $c$ values. Use
`sac_calibrate(reliability_metric = "msem")` for direct MSEM targeting.

### Jensen's Gap in Practice

```{r metric-demo}
# Calibrate with EQC under the supported info metric
eqc_info <- eqc_calibrate(
  target_rho = 0.80, n_items = 25, model = "rasch",
  item_source = "parametric", reliability_metric = "info",
  M = 5000L, seed = 42, verbose = FALSE
)

cat("Calibration with target rho* = 0.80:\n")
cat(sprintf("  Info metric: c* = %.4f, achieved = %.4f\n",
            eqc_info$c_star, eqc_info$achieved_rho))
```

```{r metric-cross-eval}
# Cross-evaluate: what does the EQC c* imply for both metrics?
both_at_info <- compute_rho_both(
  eqc_info$c_star,
  eqc_info$theta_quad,
  eqc_info$beta_vec,
  eqc_info$lambda_base,
  theta_var = eqc_info$theta_var
)

cat("\nCross-evaluation of the EQC c*:\n")
cat(sprintf("  At c*_info = %.4f: rho_tilde = %.4f, rho_bar = %.4f\n",
            eqc_info$c_star, both_at_info$rho_tilde, both_at_info$rho_bar))
cat(sprintf("  Jensen gap: %.4f\n", both_at_info$rho_tilde - both_at_info$rho_bar))
```

### 3PL calibration contract

For `model = "3pl"`, specify lower asymptotes through
`item_params$guessing_params` (or `custom_params$guessing` for a custom bank).
At every candidate scale EQC holds difficulty and guessing fixed and replaces
$\lambda_i^{(0)}$ by $c\lambda_i^{(0)}$. Setting every guessing value to zero
reproduces the 2PL kernel and reliability calculation exactly. Positive
guessing generally requires a larger selected discrimination scale for the
same fixed-form target, but this is an empirical design comparison rather than
a universal root-ordering theorem.

The symbol `c` in results and plots always denotes this global scale. It never
denotes the conventional 3PL lower asymptote; documentation and output use
`guessing` for the latter.


## Diagnostic Examples

### Monte Carlo Convergence Study

How does the EQC solution improve with quadrature size $M$?

```{r mc-convergence}
M_values <- c(500, 1000, 2000, 5000)
n_reps <- 10
target <- 0.80

# Run multiple replications at each M
mc_results <- data.frame(
  M = integer(), rep = integer(), c_star = numeric()
)

for (M_val in M_values) {
  for (r in 1:n_reps) {
    res <- eqc_calibrate(
      target_rho = target, n_items = 25, model = "rasch",
      item_source = "parametric", reliability_metric = "info",
      M = as.integer(M_val), seed = 100 * r + M_val, verbose = FALSE
    )
    mc_results <- rbind(mc_results, data.frame(
      M = M_val, rep = r, c_star = res$c_star
    ))
  }
}

# Summary statistics
mc_summary <- aggregate(c_star ~ M, data = mc_results, FUN = function(x) {
  c(mean = mean(x), sd = sd(x), min = min(x), max = max(x))
})

cat("Monte Carlo convergence of c* (target = 0.80, 25 Rasch items):\n")
cat(sprintf("  %-8s %-10s %-10s %-10s\n", "M", "Mean c*", "SD(c*)", "Range"))
for (i in seq_len(nrow(mc_summary))) {
  vals <- mc_summary$c_star[i, ]
  cat(sprintf("  %-8d %-10.4f %-10.4f [%.4f, %.4f]\n",
              mc_summary$M[i], vals["mean"], vals["sd"],
              vals["min"], vals["max"]))
}
```

```{r mc-convergence-plot, fig.cap = "Variability of EQC estimates decreases with quadrature size.", fig.alt = "Boxplots showing smaller spread in calibrated scaling factors for larger quadrature sample sizes."}
oldpar <- par(mar = c(4.5, 4.5, 3, 1))
on.exit(par(oldpar))

# Box plot of c* across M values
M_factor <- factor(mc_results$M,
                   labels = paste0("M=", format(M_values, big.mark = ",")))
boxplot(c_star ~ M_factor, data = mc_results,
        col = "lightblue", border = "steelblue",
        xlab = "Quadrature Size M", ylab = "EQC c* estimate",
        main = "Monte Carlo Variability of EQC")

# Reference line at the grand mean
abline(h = mean(mc_results$c_star[mc_results$M == max(M_values)]),
       lty = 2, col = "red", lwd = 1.5)
legend("topright", legend = "Reference (largest M)",
       lty = 2, col = "red", lwd = 1.5, cex = 0.9)
```

The standard deviation of $\hat{c}^*_M$ decreases approximately as
$1/\sqrt{M}$, consistent with the theoretical prediction from Theorem A.2.

### Sensitivity to Target Reliability

```{r target-sensitivity}
targets <- seq(0.50, 0.90, by = 0.05)
sensitivity <- data.frame(
  target  = targets,
  c_star  = numeric(length(targets)),
  achieved = numeric(length(targets))
)

for (j in seq_along(targets)) {
  res <- eqc_calibrate(
    target_rho = targets[j], n_items = 25, model = "rasch",
    item_source = "parametric", reliability_metric = "info",
    M = 5000L, seed = 42, verbose = FALSE
  )
  sensitivity$c_star[j]  <- res$c_star
  sensitivity$achieved[j] <- res$achieved_rho
}

cat("EQC sensitivity to target reliability (25 Rasch items, info metric):\n")
cat(sprintf("  %-8s %-10s %-10s %-10s\n",
            "Target", "c*", "Achieved", "|Error|"))
for (j in seq_len(nrow(sensitivity))) {
  cat(sprintf("  %-8.2f %-10.4f %-10.4f %-10.6f\n",
              sensitivity$target[j],
              sensitivity$c_star[j],
              sensitivity$achieved[j],
              abs(sensitivity$achieved[j] - sensitivity$target[j])))
}
```

```{r target-sensitivity-plot, fig.cap = "Calibrated scaling factor as a function of target reliability.", fig.alt = "Line plot showing the calibrated scaling factor increasing with the target reliability."}
oldpar <- par(mar = c(4.5, 4.5, 3, 1))
on.exit(par(oldpar))

plot(sensitivity$target, sensitivity$c_star, type = "b",
     pch = 19, col = "steelblue", lwd = 2,
     xlab = expression("Target reliability " * rho * "*"),
     ylab = expression("Calibrated scaling factor c*"),
     main = "EQC: Target vs Calibrated Scale")
grid(col = "gray90")
```

### Sensitivity to Latent Distribution Shape

```{r shape-sensitivity}
shapes <- c("normal", "bimodal", "heavy_tail", "skew_pos")
shape_pars <- list(
  normal     = list(),
  bimodal    = list(delta = 0.9),
  heavy_tail = list(df = 5),
  skew_pos   = list(k = 4)
)

cat("EQC c* for different latent shapes (target = 0.80, 25 Rasch items):\n")
for (sh in shapes) {
  res <- eqc_calibrate(
    target_rho = 0.80, n_items = 25, model = "rasch",
    item_source = "parametric", reliability_metric = "info",
    latent_shape = sh, latent_params = shape_pars[[sh]],
    M = 5000L, seed = 42, verbose = FALSE
  )
  cat(sprintf("  %-12s: c* = %.4f, achieved = %.4f\n",
              sh, res$c_star, res$achieved_rho))
}
```

### Model Comparison: Rasch vs 2PL

```{r model-comparison}
eqc_rasch <- eqc_calibrate(
  target_rho = 0.80, n_items = 25, model = "rasch",
  item_source = "parametric", reliability_metric = "info",
  M = 5000L, seed = 42, verbose = FALSE
)

eqc_2pl <- eqc_calibrate(
  target_rho = 0.80, n_items = 25, model = "2pl",
  item_source = "parametric", reliability_metric = "info",
  M = 5000L, seed = 42, verbose = FALSE
)

cat("Model comparison (target = 0.80, 25 items, info metric):\n")
cat(sprintf("  Rasch: c* = %.4f, achieved = %.4f\n",
            eqc_rasch$c_star, eqc_rasch$achieved_rho))
cat(sprintf("  2PL:   c* = %.4f, achieved = %.4f\n",
            eqc_2pl$c_star, eqc_2pl$achieved_rho))
cat(sprintf("  Rasch baseline lambda: %s\n",
            paste(unique(round(eqc_rasch$lambda_base, 2)), collapse = ", ")))
cat(sprintf("  2PL baseline lambda range: [%.2f, %.2f]\n",
            min(eqc_2pl$lambda_base), max(eqc_2pl$lambda_base)))
```

### Verbose Output Walkthrough

The verbose mode reveals the internal steps of EQC.

```{r verbose-demo}
eqc_v <- eqc_calibrate(
  target_rho = 0.80, n_items = 25, model = "rasch",
  item_source = "parametric", reliability_metric = "info",
  M = 5000L, seed = 42, verbose = TRUE
)
```

Key information from verbose output:

- **Observed range**: Reports the scanned reliability range inside `c_bounds`.
- **Root topology**: Reports detected root count, selected kind/direction, and
  scanner resolution status.
- **Achieved reliability**: Final reliability at the calibrated $c^*$.


## Output Structure

The `eqc_calibrate()` function returns an object of class `"eqc_result"`
containing the calibration results.

```{r output-structure}
# Key components
cat("EQC result components:\n")
cat(sprintf("  c_star:       %.4f  (calibrated scaling factor)\n",
            eqc_v$c_star))
cat(sprintf("  target_rho:   %.2f  (target reliability)\n",
            eqc_v$target_rho))
cat(sprintf("  achieved_rho: %.4f  (achieved reliability)\n",
            eqc_v$achieved_rho))
cat(sprintf("  metric:       %s   (reliability metric used)\n",
            eqc_v$metric))
cat(sprintf("  n_items:      %d    (number of items)\n",
            length(eqc_v$beta_vec)))
cat(sprintf("  M (quadrature): %d\n", length(eqc_v$theta_quad)))
```

### Prediction and Downstream Use

The `predict()` method evaluates reliability at arbitrary scaling factors
using the stored quadrature samples.

```{r predict-demo}
# Evaluate reliability at different c values
c_query <- c(0.5, 1.0, 1.5, 2.0)
pred <- predict(eqc_v, newdata = c_query)

cat("Predict method: reliability at different c values\n")
for (j in seq_along(c_query)) {
  cat(sprintf("  c = %.1f: rho = %.4f\n", c_query[j], pred[j]))
}
```


## Connection to SAC Validation

For rigorous validation, EQC results should be cross-checked against the
SAC algorithm using the same reliability estimand. The recommended workflow is:

1. Run EQC as the primary calibration.
2. Initialize SAC with the EQC result (warm start) and set
   `reliability_metric = "info"`.
3. Compare using `compare_eqc_sac()`.

See `vignette("algorithm-sac")` for the SAC algorithm details and
`vignette("validation")` for the complete validation framework.

```{r eqc-sac-preview}
# Quick cross-validation example
sac_check <- sac_calibrate(
  target_rho = 0.80, n_items = 25, model = "rasch",
  reliability_metric = "info", resample_items = FALSE,
  c_init = eqc_v, n_iter = 200L, M_per_iter = 1000L,
  seed = 42, verbose = FALSE,
  preflight_controls = list(M = 500L),
  evaluation_controls = list(n_forms = 3L, M = 1000L)
)

comparison <- compare_eqc_sac(eqc_v, sac_check, verbose = FALSE)
cat(sprintf("EQC-SAC agreement: %.2f%% difference\n", comparison$diff_pct))
```


## Summary

EQC is the recommended primary algorithm for reliability-targeted IRT
simulation:

| Aspect | Summary |
|:-------|:--------|
| **Method** | Adaptive log-scale topology scan plus local Brent polishing |
| **Convergence** | Root polishing is superlinear; statistical consistency is branch-conditional |
| **Asymptotic rate** | $\sqrt{M}$-normal (Theorem A.2) |
| **Supported target metric** | Average-information (`"info"` / `"tilde"`) |
| **Accuracy** | Depends on quadrature size and local slope; use independent holdout evaluation |
| **Recommended $M$** | 10,000 default; justify changes with sensitivity evidence |
| **Validation** | Cross-check with SAC via `compare_eqc_sac()` |


## References

Lee, J.-H. (2026). Reliability-Targeted Simulation of Item Response Data:
Solving the Inverse Design Problem. arXiv:2512.16012v2.
https://doi.org/10.48550/arXiv.2512.16012

Brent, R. P. (1973). *Algorithms for Minimization Without Derivatives*.
Prentice-Hall.

van der Vaart, A. W. (1998). *Asymptotic Statistics*. Cambridge University
Press.
