---
title: "Applied Guide: Reliability-Targeted IRT Simulation"
author: "JoonHo Lee (jlee296@ua.edu)"
date: "`r Sys.Date()`"
description: >
  A comprehensive practical workflow guide for reliability-targeted IRT simulation
  using IRTsimrel, covering Rasch, 2PL, and 3PL designs from feasibility
  screening through response generation and validation.
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Applied Guide: Reliability-Targeted IRT Simulation}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE, comment = "#>",
  fig.width = 7, fig.height = 5,
  fig.align = "center", out.width = "85%"
)
library(IRTsimrel)
set.seed(42)
```

# 1. Overview

**Reading time**: approximately 30--40 minutes.

This vignette walks through the complete workflow for generating simulated item
response data with a **pre-specified marginal reliability**. The approach
implements the framework described in Lee (2026), which treats reliability as a
design parameter rather than an emergent property of the simulation.

**Learning objectives.** After completing this guide you will be able to:

1. Screen whether a desired reliability is achievable for a given test design.
2. Visualize the reliability curve to understand the design space.
3. Calibrate a discrimination scaling factor with EQC (Algorithm 1).
4. Optionally validate the calibration with SAC (Algorithm 2).
5. Generate item response data at the target reliability.
6. Extract, summarize, and export calibrated item parameters.
7. Distinguish fixed-form from item-superpopulation targets in 3PL work.

**Prerequisites.** Install IRTsimrel and, optionally, TAM for external
Rasch/2PL validation and 3PL EAP diagnostics:

```{r install-packages, eval = FALSE}
# install.packages("devtools")
devtools::install_github("joonho112/IRTsimrel")
install.packages("TAM")
```

In the TAM version used for the package's v0.3 validation, a 3PL fit supports
EAP scoring but not `tam.wle()` on the `tam.mml.3pl` object. Do not coerce the
object class or report a 3PL TAM WLE result through an unsupported path.

Most baseline examples use `seed = 42` and `M = 5000L`; the compact 3PL
workflow uses smaller documented settings and separate seeds so the vignette
remains quick and reproducible. Treat these as tutorial settings, not automatic
precision recommendations for a final study.


# 2. The 6-Step Workflow

The reliability-targeted simulation workflow consists of six steps.  The table
below provides a quick summary before we walk through each step in detail.

| Step | Function | Purpose |
|:----:|:---------|:--------|
| 1 | `check_feasibility()` | Screen whether $\rho^*$ is achievable |
| 2 | `rho_curve()` | Visualize $\rho(c)$ across scaling factors |
| 3 | `eqc_calibrate()` | Find $c^*$ such that $\rho(c^*) \approx \rho^*$ |
| 4 | `sac_calibrate()` | (Optional) Validate $c^*$ via stochastic approximation |
| 5 | `simulate_response_data()` | Generate binary response matrix |
| 6 | `coef()`, `predict()`, `as.data.frame()` | Extract and use results |


## 2.1 Step 1: Screen Feasibility with `check_feasibility()`

Before investing computation time in calibration, verify that the target
reliability is achievable.  `check_feasibility()` evaluates both reliability
metrics across a range of scaling factors and reports the achievable interval.

```{r step1-feasibility}
feas <- check_feasibility(
  n_items      = 25,
  model        = "rasch",
  latent_shape = "normal",
  item_source  = "parametric",
  target_rho   = 0.85,
  c_bounds     = c(0.1, 10),
  M            = 5000L,
  seed         = 42
)
```

The output reports the resolved range for both the average-information metric
($\tilde{\rho}$) and the MSEM-based metric ($\bar{w}$) in one finite Monte
Carlo design. If a target falls outside that range, it is infeasible within
the configured bounds and sampled design. Inspect the topology and
best-achievable record before deciding whether to justify wider bounds, add
items, or change the item model. A finite heavy-tail MSEM range is sensitivity
evidence only, not a population feasibility certificate.

**Quick check for a specific target:**

```{r step1-check-target}
target <- 0.85
in_range <- identical(feas$target_status_info_canonical, "feasible")
admissible <- feas$admissible_root_count_info > 0L
cat(sprintf("Target rho = %.2f is %s for this design.\n",
            target,
            ifelse(in_range && admissible,
                   "FEASIBLE WITH AN ADMISSIBLE ROOT",
                   "NOT YET CALIBRATION-READY")))
```

### Comparing feasibility across designs

You can compare different test lengths to find the minimum number of items
needed:

```{r step1-compare-designs}
test_lengths <- c(10, 15, 20, 25, 30, 40)
feas_table <- data.frame(
  n_items   = test_lengths,
  rho_min   = numeric(length(test_lengths)),
  rho_max   = numeric(length(test_lengths))
)

for (i in seq_along(test_lengths)) {
  fi <- check_feasibility(
    n_items = test_lengths[i], model = "rasch",
    item_source = "parametric", target_rho = target, M = 5000L,
    seed = 42, verbose = FALSE
  )
  feas_table$rho_min[i] <- round(fi$rho_range_info[1], 4)
  feas_table$rho_max[i] <- round(fi$rho_range_info[2], 4)
}

feas_table
```

```{r step1-feas-plot, fig.height = 4, fig.alt = "Line plot of achievable reliability ranges by test length with the target reliability shown as a horizontal reference line."}
plot(NULL, xlim = range(test_lengths), ylim = c(0, 1),
     xlab = "Number of Items", ylab = expression(tilde(rho)),
     main = "Achievable Reliability Range by Test Length")
polygon(c(test_lengths, rev(test_lengths)),
        c(feas_table$rho_min, rev(feas_table$rho_max)),
        col = adjustcolor("steelblue", alpha.f = 0.3), border = NA)
lines(test_lengths, feas_table$rho_min, col = "steelblue", lwd = 2)
lines(test_lengths, feas_table$rho_max, col = "steelblue", lwd = 2)
abline(h = 0.85, lty = 2, col = "firebrick")
text(max(test_lengths), 0.86, expression(rho^"*" == 0.85),
     col = "firebrick", pos = 2, cex = 0.9)
```

The shaded region shows the achievable reliability range as a function of test
length.  This type of visualization is useful for study planning: you can
immediately see the minimum test length needed to reach your desired reliability.

For more on how the latent distribution affects these ranges, see
`vignette("latent-distributions")`.


## 2.2 Step 2: Explore the Reliability Curve with `rho_curve()`

The reliability curve $\rho(c)$ shows how reliability changes as the global
discrimination scaling factor $c$ varies.  This visualization helps you
understand the shape of the design space and select sensible targets.

```{r step2-rho-curve, fig.height = 5, fig.alt = "Reliability curve showing how reliability changes with the discrimination scaling factor for the selected design."}
curve_data <- rho_curve(
  n_items      = 25,
  model        = "rasch",
  latent_shape = "normal",
  item_source  = "parametric",
  metric       = "both",
  M            = 5000L,
  seed         = 42,
  plot         = TRUE
)
```

The plot shows both $\tilde{\rho}$ (average-information) and $\bar{w}$
(MSEM-based).  Note that $\tilde{\rho} \geq \bar{w}$ holds on the same
population/grid basis by Jensen's inequality.

**Reading the curve.** You can use the returned data frame to find approximate
$c$ values for any target:

```{r step2-read-curve}
head(curve_data)

# Find approximate c for rho_tilde = 0.80
idx <- which.min(abs(curve_data$rho_tilde - 0.80))
cat(sprintf("Approximate c for rho_tilde = 0.80: %.3f\n", curve_data$c[idx]))
```


## 2.3 Step 3: Calibrate with EQC --- `eqc_calibrate()`

EQC (Empirical Quadrature Calibration) is Algorithm 1 from Lee (2026). It maps
the empirical reliability topology on the requested log-scale interval,
selects a branch under `root_policy`, and uses Brent's method via
`uniroot()` only to polish ordinary sign-changing crossings of
$\hat{\rho}_M(c^*) = \rho^*$.

### 2.3.1 Basic calibration (Rasch)

```{r step3-eqc-basic}
eqc_result <- eqc_calibrate(
  target_rho        = 0.85,
  n_items           = 25,
  model             = "rasch",
  latent_shape      = "normal",
  item_source       = "parametric",
  reliability_metric = "info",
  M                 = 5000L,
  seed              = 42,
  verbose           = TRUE
)
```

```{r step3-eqc-print}
print(eqc_result)
```

**Key outputs:**

- `c_star`: the calibrated scaling factor $c^*$.
- `achieved_rho`: the empirical reliability at $c^*$.
- `items_calib`: the item parameter object with scaled discriminations.

### 2.3.2 Exploring the result

```{r step3-summary}
summary(eqc_result)
```

### 2.3.3 Key parameters explained

| Parameter | Default | Description |
|:----------|:--------|:------------|
| `target_rho` | --- | Target marginal reliability $\rho^* \in (0, 1)$ |
| `n_items` | --- | Number of items on the test form |
| `model` | `"rasch"` | `"rasch"`, `"2pl"`, or `"3pl"` |
| `latent_shape` | `"normal"` | Latent distribution shape (see `vignette("latent-distributions")`) |
| `item_source` | `"parametric"` | Item parameter source (see `vignette("item-parameters")`) |
| `reliability_metric` | `"info"` | EQC supports `"info"` / `"tilde"`; use SAC for direct `"msem"` targeting |
| `M` | `10000L` | Monte Carlo sample size for quadrature |
| `c_bounds` | `c(0.3, 3)` | Search interval for $c$ |
| `seed` | `NULL` | For reproducibility |

### 2.3.4 Calibrating under different models and shapes

**2PL model with bimodal population:**

```{r step3-eqc-2pl-bimodal}
eqc_2pl <- eqc_calibrate(
  target_rho   = 0.80,
  n_items      = 30,
  model        = "2pl",
  latent_shape = "bimodal",
  latent_params = list(shape_params = list(delta = 0.8)),
  item_source  = "parametric",
  reliability_metric = "info",
  M            = 5000L,
  seed         = 42
)

cat(sprintf("c* = %.4f, achieved rho = %.4f\n",
            eqc_2pl$c_star, eqc_2pl$achieved_rho))
```

**Skewed population with heavy tails:**

```{r step3-eqc-skew}
eqc_skew <- eqc_calibrate(
  target_rho   = 0.80,
  n_items      = 25,
  model        = "rasch",
  latent_shape = "skew_pos",
  item_source  = "parametric",
  reliability_metric = "info",
  M            = 5000L,
  seed         = 42
)

cat(sprintf("c* = %.4f, achieved rho = %.4f\n",
            eqc_skew$c_star, eqc_skew$achieved_rho))
```


## 2.4 Step 4: (Optional) Validate with SAC --- `sac_calibrate()`

SAC (Stochastic Approximation Calibration) is Algorithm 2.  It uses the
Robbins--Monro stochastic approximation algorithm and can target either
$\tilde{\rho}$ or $\bar{w}$.  SAC is useful for:

- **Independent validation** of EQC results.
- **Targeting $\bar{w}$** directly (the MSEM-based reciprocal-information metric).
- **Item-superpopulation calibration** when forms, rather than only persons,
  are part of the target distribution.

### 2.4.1 Warm-start from EQC (recommended)

Passing the EQC result as `c_init` accelerates SAC convergence:

```{r step4-sac-warm}
sac_result <- sac_calibrate(
  target_rho        = 0.85,
  n_items           = 25,
  model             = "rasch",
  latent_shape      = "normal",
  reliability_metric = "info",
  c_init            = eqc_result,
  resample_items    = FALSE,
  n_iter            = 200L,
  M_per_iter        = 500L,
  M_pre             = 5000L,
  seed              = 42,
  verbose           = TRUE
)
```

### 2.4.2 Visualize SAC convergence

```{r step4-sac-plot, fig.height = 4, fig.alt = "SAC convergence diagnostic plot for the warm-start validation run."}
plot(sac_result)
```

### 2.4.3 Compare EQC and SAC

```{r step4-compare}
comparison <- compare_eqc_sac(eqc_result, sac_result)
```

The comparison reports the absolute and percentage difference in $c^*$ between
the two algorithms and applies a 5% diagnostic threshold only after verifying
that their estimand and item-design contracts are comparable.

### 2.4.4 SAC parameter reference

| Parameter | Default | Description |
|:----------|:--------|:------------|
| `c_init` | `NULL` (APC) | Initial $c_0$; pass an `eqc_result` for warm start |
| `n_iter` | `300L` | Total Robbins--Monro iterations |
| `M_per_iter` | `500L` | MC samples per iteration |
| `M_pre` | `10000L` | MC samples for pre-calculating $\sigma^2_\theta$ |
| `burn_in` | `floor(n_iter/2)` | Iterations to discard before averaging |
| `step_params` | `list(a=1, A=50, gamma=0.67)` | Step-size sequence: $a_n = a/(n+A)^\gamma$ |
| `resample_items` | `TRUE` | `TRUE`: item-superpopulation target; `FALSE`: one fixed form |

EQC always calibrates the one generated form. SAC has two explicit scopes:

- With `resample_items = FALSE`, SAC reuses one form. When `c_init` is an EQC
  object and `item_source`/`item_params` are omitted, it reuses the exact EQC
  bank and verifies the same target, model, metric, item count, and form.
- With `resample_items = TRUE`, SAC redraws forms and reports the mean
  reliability over independent holdout forms. This is an item-superpopulation
  estimand, so an EQC object is not a comparable warm start; use the automatic
  initializer or a numeric starting value.


## 2.5 Step 5: Generate Response Data --- `simulate_response_data()`

Once calibration is complete, generate a binary response matrix.  The function
draws $N$ persons from the specified latent distribution and generates responses
using the calibrated item parameters.

```{r step5-sim-data}
sim_data <- simulate_response_data(
  result       = eqc_result,
  n_persons    = 500,
  latent_shape = "normal",
  seed         = 123
)

cat(sprintf("Response matrix: %d persons x %d items\n",
            nrow(sim_data$response_matrix),
            ncol(sim_data$response_matrix)))
cat(sprintf("Mean proportion correct: %.3f\n",
            mean(sim_data$response_matrix)))
sim_data$provenance[c("metric", "calibration_status", "status_flags",
                      "item_source", "simulation_seed", "latent_shape")]
```

### 2.5.1 Inspect the generated data

```{r step5-inspect}
# First 5 persons, first 5 items
sim_data$response_matrix[1:5, 1:5]

# Distribution of total scores
total_scores <- rowSums(sim_data$response_matrix)
summary(total_scores)
```

```{r step5-score-hist, fig.height = 4, fig.alt = "Histogram of total scores from the simulated response matrix."}
hist(total_scores, breaks = 20, col = "steelblue",
     border = "white", main = "Distribution of Total Scores",
     xlab = "Total Score", ylab = "Frequency")
```

### 2.5.2 Classical item analysis

A quick sanity check: compute item-total correlations and item difficulty
(proportion correct) from the generated data.

```{r step5-item-analysis, fig.height = 4.5, fig.alt = "Scatter plot of item difficulty against observed proportion correct in the simulated data."}
p_correct <- colMeans(sim_data$response_matrix)
item_total_cor <- apply(sim_data$response_matrix, 2, function(x) {
  cor(x, total_scores - x)  # corrected item-total correlation
})

par(mfrow = c(1, 2))
hist(p_correct, breaks = 15, col = "steelblue", border = "white",
     main = "Item Difficulty (p)", xlab = "Proportion Correct")
hist(item_total_cor, breaks = 15, col = "darkorange", border = "white",
     main = "Item-Total Correlation", xlab = "Corrected r_it")
par(mfrow = c(1, 1))
```

### 2.5.3 External validation with TAM

If you have the TAM package installed, you can verify that the achieved
reliability provides a useful external diagnostic for Rasch/2PL data:

```{r step5-tam-validation, eval = FALSE}
if (requireNamespace("TAM", quietly = TRUE)) {
  tam_rel <- compute_reliability_tam(sim_data$response_matrix, model = "rasch")
  cat(sprintf("Target reliability: %.4f\n", eqc_result$target_rho))
  cat(sprintf("EQC achieved rho:   %.4f\n", eqc_result$achieved_rho))
  cat(sprintf("TAM WLE reliability: %.4f\n", tam_rel$rel_wle))
  cat(sprintf("TAM EAP reliability: %.4f\n", tam_rel$rel_eap))
}
```

> **Note on WLE vs EAP reliability.** In many practical TAM fits, EAP
> reliability is higher than WLE reliability, but `EAPrel()` and `WLErel()` use
> different estimators and variance bases. Inspect both as complementary
> external diagnostics rather than formal lower/upper bounds.

For 3PL data, fit `TAM::tam.mml.3pl()` directly and use its EAP person output
or `EAP.rel` only. `compute_reliability_tam()` intentionally exposes the
Rasch/2PL helper contract, and `TAM::tam.wle()` was unsupported for a
`tam.mml.3pl` object in the validated TAM version. Moreover, EAP score
reliability is estimator-, sample-, prior-, and fit-dependent; it is not the
same estimand as IRTsimrel's analytic population information reliability.


## 2.6 Step 6: Extract and Use Results

### 2.6.1 `coef()` --- calibrated item parameters

The `coef()` method extracts a tidy data frame of item parameters:

```{r step6-coef}
item_df <- coef(eqc_result)
head(item_df)
```

Each row contains:

- `item_id`: item identifier.
- `beta`: item difficulty.
- `lambda_base`: baseline (unscaled) discrimination.
- `lambda_scaled`: calibrated discrimination ($\lambda_{\text{base}} \times c^*$).
- `c_star`: the calibrated scaling factor (constant across items).

### 2.6.2 `predict()` --- reliability at new scaling factors

Use `predict()` to evaluate reliability at arbitrary scaling factor values:

```{r step6-predict}
# Achieved reliability
predict(eqc_result)

# Reliability at several c values
predict(eqc_result, newdata = c(0.5, 1.0, 1.5, 2.0))
```

This is useful for sensitivity analysis---for instance, exploring how much
reliability changes if the scaling factor shifts by $\pm 10\%$.

### 2.6.3 `as.data.frame()` --- export item parameters

For downstream analysis, export the calibrated item parameters to a data frame:

```{r step6-as-df}
items_df <- as.data.frame(eqc_result$items_calib)
head(items_df)
```

### 2.6.4 Accessing internal components

The calibration result objects store additional information for advanced use:

```{r step6-components}
# EQC components
cat("EQC result components:\n")
cat(paste(" ", names(eqc_result), collapse = "\n"), "\n\n")

# Calibrated scaling factor
cat(sprintf("c* = %.4f\n", eqc_result$c_star))

# Latent variance from quadrature sample
cat(sprintf("theta_var = %.4f\n", eqc_result$theta_var))
```


## 2.7 Compact End-to-End 3PL Workflow

A 3PL form has lower asymptotes $g_i$ in addition to difficulty and
discrimination:

$$
P(Y_i=1\mid\theta)=g_i+(1-g_i)
\operatorname{logit}^{-1}\{c\lambda_{i,0}(\theta-\beta_i)\}.
$$

`guessing_params` controls $g_i$. The global scale $c$ multiplies only the
baseline discriminations $\lambda_{i,0}$; it never multiplies the guessing
parameters. The next executable example makes that separation explicit and
freezes the generated form before calibration.

```{r workflow-3pl-generate}
generated_3pl <- sim_item_params(
  n_items = 12,
  model = "3pl",
  source = "parametric",
  method = "copula",
  guessing_params = list(distribution = "fixed", value = 0.20),
  scale = 1,
  seed = 310
)

fixed_3pl_spec <- list(custom_params = list(
  beta = generated_3pl$data$beta,
  lambda = generated_3pl$data$lambda_unscaled,
  guessing = generated_3pl$data$guessing
))
```

EQC estimates a discrimination scale for that fixed form. The smaller
quadrature setting below is suitable for documentation; increase `M` and use
independent seeded holdouts for a final study.

```{r workflow-3pl-eqc}
eqc_3pl <- eqc_calibrate(
  target_rho = 0.70,
  n_items = 12,
  model = "3pl",
  item_source = "custom",
  item_params = fixed_3pl_spec,
  reliability_metric = "info",
  M = 2000L,
  seed = 310
)

coef(eqc_3pl)[1:6, c(
  "beta", "lambda_base", "lambda_scaled", "guessing", "c_star"
)]
```

For a same-estimand stochastic check, set `resample_items = FALSE` and omit
item-generation arguments. SAC then reuses the exact EQC bank and rejects a
mismatched metric or form instead of silently comparing unlike designs.

```{r workflow-3pl-sac}
sac_3pl <- sac_calibrate(
  target_rho = 0.70,
  n_items = 12,
  model = "3pl",
  reliability_metric = "info",
  c_init = eqc_3pl,
  resample_items = FALSE,
  n_iter = 60L,
  M_per_iter = 250L,
  M_pre = 1000L,
  evaluation_controls = list(n_forms = 3L, M = 500L),
  seed = 310
)

same_form_check <- compare_eqc_sac(eqc_3pl, sac_3pl)
stopifnot(same_form_check$comparable)
```

Finally, generate responses from the calibrated form and verify its structural
contract. `simulate_response_data()` preserves the calibrated discriminations,
the fixed lower asymptotes, and provenance.

```{r workflow-3pl-response}
sim_3pl <- simulate_response_data(
  result = eqc_3pl,
  n_persons = 200,
  seed = 311
)

stopifnot(
  identical(dim(sim_3pl$response_matrix), c(200L, 12L)),
  all(sim_3pl$response_matrix %in% 0:1),
  all(sim_3pl$guessing == 0.20),
  identical(sim_3pl$provenance$item_scope, "fixed_form")
)
c(
  eqc_rho = eqc_3pl$achieved_rho,
  sac_holdout_rho = sac_3pl$achieved_rho,
  response_rate = mean(sim_3pl$response_matrix)
)
```

For an item-superpopulation target, use SAC without an EQC-object warm start:

```{r workflow-3pl-superpopulation, eval = FALSE}
sac_3pl_population <- sac_calibrate(
  target_rho = 0.70,
  n_items = 12,
  model = "3pl",
  item_params = list(
    guessing_params = list(distribution = "fixed", value = 0.20)
  ),
  reliability_metric = "info",
  c_init = NULL,
  resample_items = TRUE,
  n_iter = 300L,
  M_per_iter = 1000L,
  M_pre = 10000L,
  seed = 310
)
```

Its `achieved_rho` is the mean over independent holdout-form reliabilities,
not the reliability of the single representative form stored for convenient
response generation. Setting every $g_i=0$ recovers exact 2PL behavior in the
shared probability and information kernels; existing Rasch/2PL calls remain
valid without any new arguments.


# 3. Choosing EQC vs SAC

The following decision table helps you choose between the two algorithms.

| Criterion | EQC (Algorithm 1) | SAC (Algorithm 2) |
|:----------|:-------------------|:-------------------|
| **Speed** | Fast (adaptive topology scan plus local root polishing) | Slower (preflight plus iterative updates) |
| **Default metric** | $\tilde{\rho}$ (info) | $\bar{w}$ (msem) |
| **Deterministic?** | Conditional on the frozen Monte Carlo design | No (stochastic) |
| **Topology contract** | Maps the full requested log-scale interval, enumerates crossings/extrema, and applies `root_policy`; no global monotonicity assumption | Preflight must resolve a feasible increasing branch before updates |
| **Can target $\bar{w}$?** | No; use `"info"` / `"tilde"` | Yes, natively |
| **Warm start** | Not needed | Compatible EQC object for a same-metric fixed-form run; numeric scale otherwise |
| **Use for validation** | Primary calibration | Same-estimand check with `reliability_metric = "info"` |
| **Diagnostics** | Root topology, selected branch, and empirical residual | Trajectory, branch, projection, holdout, and Polyak-average diagnostics |

**Practical recommendation:**

1. **Start with EQC** using `reliability_metric = "info"`. This is fast and
   reproducible with a fixed seed and stored form/quadrature design.
2. **Use SAC with `reliability_metric = "info"` to validate EQC** when you want
   a stochastic same-estimand check.
3. **Use SAC with `reliability_metric = "msem"` separately** when your research
   question targets $\bar{w}$ directly.
4. **Warm-start fixed-form SAC** from EQC when comparing the same form and
   metric. Do not use an EQC-object warm start for an item-superpopulation run.


# 4. Choosing the Reliability Metric: info vs msem

IRTsimrel supports two population reliability definitions.

## 4.1 Average-information reliability (rho-tilde) {#average-information-reliability-tilderho}

$$
\tilde{\rho}(c) = \frac{\sigma^2_\theta \, \bar{\mathcal{J}}(c)}
{\sigma^2_\theta \, \bar{\mathcal{J}}(c) + 1}
$$

where $\bar{\mathcal{J}}(c) = \mathbb{E}_G[\mathcal{J}(\theta; c)]$ is the
average test information across the latent distribution.

**Properties:**

- Often well behaved on practical intervals, but saturation can create an
  interior maximum, multiple crossings, or boundary-only feasibility. EQC maps
  the requested interval adaptively and selects among detected roots with an
  explicit `root_policy`.
- Upper bound on $\bar{w}$ via Jensen's inequality.
- Faster to compute than $\bar{w}$.

## 4.2 MSEM-based marginal reliability (w-bar) {#msem-based-marginal-reliability-barw}

$$
\bar{w}(c) = \frac{\sigma^2_\theta}
{\sigma^2_\theta + \mathbb{E}_G[1/\mathcal{J}(\theta; c)]}
$$

**Properties:**

- MSEM-based reliability using the reciprocal-information approximation to
  conditional error variance.
- Can be non-monotone in $c$ for extreme scaling.
- Requires SAC for safe targeting.

## 4.3 Jensen's inequality: the key relationship

By Jensen's inequality applied to the convex function $f(x) = 1/x$:

$$
\mathbb{E}\!\left[\frac{1}{\mathcal{J}(\theta)}\right]
\;\geq\;
\frac{1}{\mathbb{E}[\mathcal{J}(\theta)]}
$$

which implies $\tilde{\rho} \geq \bar{w}$.  The gap is small when test
information is approximately constant across $\theta$, and larger when
information varies substantially (e.g., short tests, non-normal populations).

## 4.4 Decision guidance

| Scenario | Recommended metric | Reason |
|:---------|:-------------------|:-------|
| Standard simulation study | `"info"` | Stable on practical intervals; fast seeded root-finding |
| MSEM-based marginal reliability needed | `"msem"` via SAC | Uses reciprocal-information error variance |
| Non-normal latent distribution | `"info"` for EQC | Bracket-checked seeded root-finding |
| EQC/SAC validation | `"info"` for both | Compare the same estimand via `compare_eqc_sac()` |

```{r metric-comparison}
# Compare metrics for the same calibration
eqc_info <- eqc_calibrate(
  target_rho = 0.85, n_items = 25, model = "rasch",
  reliability_metric = "info", M = 5000L, seed = 42
)

cat(sprintf("Targeting info:  c* = %.4f, achieved = %.4f\n",
            eqc_info$c_star, eqc_info$achieved_rho))
```


# 5. Working with Different Models

## 5.1 Rasch model

In the Rasch model, all baseline discriminations are equal to 1.  The calibrated
discriminations are $\lambda_i^* = c^* \times 1 = c^*$ for all items.

```{r model-rasch}
eqc_rasch <- eqc_calibrate(
  target_rho  = 0.80,
  n_items     = 20,
  model       = "rasch",
  item_source = "parametric",
  reliability_metric = "info",
  M           = 5000L,
  seed        = 42
)

items_rasch <- coef(eqc_rasch)
cat(sprintf("All lambda_scaled equal? %s\n",
            all(items_rasch$lambda_scaled == items_rasch$lambda_scaled[1])))
cat(sprintf("Common discrimination: %.4f\n", items_rasch$lambda_scaled[1]))
```

## 5.2 Two-parameter logistic (2PL) model

In the 2PL model, baseline discriminations vary across items.  The scaling
factor $c^*$ is applied uniformly:

$$\lambda_i^* = c^* \times \lambda_{i,0}$$

```{r model-2pl}
eqc_2pl_demo <- eqc_calibrate(
  target_rho  = 0.80,
  n_items     = 25,
  model       = "2pl",
  item_source = "parametric",
  reliability_metric = "info",
  M           = 5000L,
  seed        = 42
)

items_2pl <- coef(eqc_2pl_demo)
cat(sprintf("c* = %.4f\n", eqc_2pl_demo$c_star))
cat(sprintf("lambda_scaled range: [%.3f, %.3f]\n",
            min(items_2pl$lambda_scaled), max(items_2pl$lambda_scaled)))
```

## 5.3 Three-parameter logistic (3PL) model

The 3PL retains the varying baseline discrimination of the 2PL and adds a
lower asymptote. Reliability calibration changes the common discrimination
scale while preserving each item's difficulty and guessing parameter.

```{r model-3pl}
eqc_3pl_demo <- eqc_calibrate(
  target_rho = 0.70,
  n_items = 12,
  model = "3pl",
  item_source = "parametric",
  item_params = list(
    guessing_params = list(distribution = "uniform", min = 0.15, max = 0.25)
  ),
  reliability_metric = "info",
  M = 2000L,
  seed = 42
)

items_3pl_demo <- coef(eqc_3pl_demo)
range(items_3pl_demo$guessing)
stopifnot(all.equal(
  items_3pl_demo$lambda_scaled,
  eqc_3pl_demo$c_star * items_3pl_demo$lambda_base
))
```

Because nonzero guessing changes the information curve topology, a 3PL
calibration can have more than one candidate crossing on a wide scale range.
IRTsimrel scans the bounded scale interval and, by default, selects the lowest
increasing root. Inspect the stored topology/status instead of assuming that
endpoint monotonicity is sufficient.

## 5.4 Rasch vs 2PL comparison

```{r model-comparison}
targets <- c(0.70, 0.75, 0.80, 0.85, 0.90)
comp_df <- data.frame(
  target  = targets,
  c_rasch = numeric(length(targets)),
  c_2pl   = numeric(length(targets))
)

for (i in seq_along(targets)) {
  r1 <- eqc_calibrate(target_rho = targets[i], n_items = 25, model = "rasch",
                       item_source = "parametric", reliability_metric = "info",
                       M = 5000L, seed = 42)
  r2 <- eqc_calibrate(target_rho = targets[i], n_items = 25, model = "2pl",
                       item_source = "parametric", reliability_metric = "info",
                       M = 5000L, seed = 42)
  comp_df$c_rasch[i] <- round(r1$c_star, 4)
  comp_df$c_2pl[i]   <- round(r2$c_star, 4)
}

comp_df
```

## 5.5 How latent shape affects calibration

Different latent distribution shapes require different scaling factors to
achieve the same target reliability, even with the same item parameters.  This
is because the test information function interacts differently with each shape.

```{r shape-sensitivity}
shapes <- c("normal", "bimodal", "skew_pos", "heavy_tail", "uniform")
shape_df <- data.frame(
  shape      = shapes,
  c_star     = numeric(length(shapes)),
  achieved   = numeric(length(shapes))
)

for (i in seq_along(shapes)) {
  ri <- eqc_calibrate(
    target_rho = 0.80, n_items = 25, model = "rasch",
    latent_shape = shapes[i], item_source = "parametric",
    reliability_metric = "info", M = 5000L, seed = 42
  )
  shape_df$c_star[i]   <- round(ri$c_star, 4)
  shape_df$achieved[i] <- round(ri$achieved_rho, 4)
}

shape_df
```

```{r shape-barplot, fig.height = 4, fig.alt = "Bar plot comparing achieved reliability across latent distribution shapes."}
barplot(
  shape_df$c_star,
  names.arg = shape_df$shape,
  col = "steelblue", border = "white",
  main = "Calibrated c* by Latent Shape (target rho = 0.80)",
  ylab = expression(c^"*"), las = 2, cex.names = 0.8
)
abline(h = mean(shape_df$c_star), lty = 2, col = "grey40")
```

The barplot reveals that heavy-tailed and skewed distributions generally require
higher scaling factors (stronger discriminations) to achieve the same
reliability.  This makes intuitive sense: extreme $\theta$ values receive less
test information under the logistic model, and heavier tails put more mass in
those regions.

## 5.6 Effect of test length on reliability curve

```{r test-length-curves, fig.height = 5, fig.alt = "Reliability curves comparing how test length affects the discrimination scale needed for a target reliability."}
lengths_to_plot <- c(10, 20, 30, 50)
cols <- c("firebrick", "darkorange", "steelblue", "forestgreen")
c_grid <- seq(0.2, 4, length.out = 40)

plot(NULL, xlim = c(0.2, 4), ylim = c(0, 1),
     xlab = "Scaling factor c", ylab = expression(tilde(rho)),
     main = "Reliability Curves by Test Length")
abline(h = 0.80, lty = 2, col = "grey50")

for (j in seq_along(lengths_to_plot)) {
  cd <- rho_curve(
    c_values = c_grid, n_items = lengths_to_plot[j],
    model = "rasch", latent_shape = "normal",
    item_source = "parametric", metric = "info",
    M = 5000L, seed = 42, plot = FALSE
  )
  lines(cd$c, cd$rho_tilde, col = cols[j], lwd = 2)
}

legend("bottomright",
       legend = paste0("I = ", lengths_to_plot),
       col = cols, lwd = 2, bty = "n")
```

The figure illustrates a key relationship: longer tests reach any given
reliability target at a lower scaling factor.  This means that adding items is
an alternative to increasing discrimination when a high reliability is needed.


# 6. Troubleshooting Guide

## 6.1 Common issues and solutions

### "Target reliability not achievable"

**Symptom:** `check_feasibility()` shows the target is outside the achievable
range.

**Solutions:**

- Increase the number of items.
- Switch from Rasch to 2PL (varying discriminations provide more flexibility).
- For 3PL designs, inspect the selected root topology and reconsider the
  guessing distribution or scale bounds rather than assuming monotonicity.
- Widen `c_bounds` in `check_feasibility()` to explore a larger range.
- Consider whether the latent shape makes the target unrealistic.

### `uniroot()` fails with "values at endpoints not of opposite sign"

**Symptom:** EQC calibration fails because the target lies outside the range
$[\rho(c_{\min}), \rho(c_{\max})]$.

**Solutions:**

- Run `check_feasibility()` first.
- Widen `c_bounds` in `eqc_calibrate()`.
- If targeting `"msem"`, switch to `"info"` (the bracket-checked EQC metric).

### SAC does not converge

**Symptom:** The SAC trajectory oscillates without settling.

**Solutions:**

- Increase `n_iter` (e.g., from 300 to 500 or more).
- Decrease the step-size base `a` in `step_params`.
- Increase `A` in `step_params` for more stabilization.
- Use a warm start from EQC.
- Increase `M_per_iter` to reduce per-iteration variance.

### Results differ between EQC and SAC

**Symptom:** `compare_eqc_sac()` reports more than 5% difference.

**Possible causes:**

- Different reliability metrics (check `metric` field).
- Different item scopes (fixed form versus item superpopulation) or item banks.
- Insufficient Monte Carlo samples (`M` for EQC, `M_per_iter`/`n_iter` for SAC).
- Different random seeds producing different item/theta draws.

**Solutions:**

- Increase `M` and `M_per_iter`.
- Use `reliability_metric = "info"` for both algorithms for fair comparison.
- For a fixed-form comparison, pass the EQC result with
  `resample_items = FALSE` and omit item-generation arguments so the same form
  is reused. A shared seed alone does not make different estimands comparable.

## 6.2 Performance tips

| Goal | Recommendation |
|:-----|:---------------|
| Fast exploration | Use `M = 5000L`, `n_iter = 100L` |
| Higher-precision starting point | Try `M = 50000L`, `n_iter = 500L`, then verify seed and holdout sensitivity |
| Quick feasibility check | Use `check_feasibility()` with `M = 3000L` |
| Reduce SAC variance | Increase `M_per_iter` to 1000 or 2000 |
| Faster SAC | Warm start from EQC, reduce `n_iter` |


# 7. Complete Template: Copy-Paste Workflow

The following code block is a self-contained template you can copy into your
project.  Replace the settings in the "Configuration" section with your own.

```{r template-workflow, eval = FALSE}
# ============================================================
# Reliability-Targeted IRT Simulation: Complete Workflow
# ============================================================
library(IRTsimrel)

# --- Configuration ------------------------------------------
target_rho   <- 0.85       # desired marginal reliability
n_items      <- 25         # test length
model        <- "rasch"    # "rasch", "2pl", or "3pl"
latent_shape <- "normal"   # latent distribution shape
item_source  <- "parametric"
item_params  <- if (model == "3pl") {
  list(guessing_params = list(distribution = "fixed", value = 0.20))
} else {
  list()
}
N_persons    <- 500        # sample size for response data
M_quad       <- 5000L      # Monte Carlo samples for EQC
seed_val     <- 42         # for reproducibility

# --- Step 1: Feasibility ------------------------------------
feas <- check_feasibility(
  n_items = n_items, model = model,
  latent_shape = latent_shape, item_source = item_source,
  item_params = item_params,
  target_rho = target_rho, M = M_quad, seed = seed_val, verbose = FALSE
)
stopifnot(
  identical(feas$target_status_info_canonical, "feasible"),
  feas$admissible_root_count_info > 0L
)
cat("Step 1: Feasible target with an admissible root confirmed.\n")

# --- Step 2: Reliability curve (optional) --------------------
curve_df <- rho_curve(
  n_items = n_items, model = model,
  latent_shape = latent_shape, item_source = item_source,
  item_params = item_params,
  M = M_quad, seed = seed_val, plot = FALSE
)

# --- Step 3: EQC calibration --------------------------------
eqc_res <- eqc_calibrate(
  target_rho = target_rho, n_items = n_items,
  model = model, latent_shape = latent_shape,
  item_source = item_source,
  item_params = item_params,
  reliability_metric = "info",
  M = M_quad, seed = seed_val
)
cat(sprintf("Step 3: EQC calibrated c* = %.4f, achieved rho = %.4f\n",
            eqc_res$c_star, eqc_res$achieved_rho))

# --- Step 4: SAC validation (optional) ----------------------
sac_res <- sac_calibrate(
  target_rho = target_rho, n_items = n_items,
  model = model, latent_shape = latent_shape,
  reliability_metric = "info",
  c_init = eqc_res,
  resample_items = FALSE,
  n_iter = 200L, M_per_iter = 500L, M_pre = 5000L,
  seed = seed_val
)
cat(sprintf("Step 4: SAC calibrated c* = %.4f\n", sac_res$c_star))

# --- Step 5: Generate response data -------------------------
sim_data <- simulate_response_data(
  result = eqc_res, n_persons = N_persons,
  latent_shape = latent_shape, seed = 123
)
cat(sprintf("Step 5: Generated %d x %d response matrix.\n",
            nrow(sim_data$response_matrix),
            ncol(sim_data$response_matrix)))

# --- Step 6: Extract results --------------------------------
item_params <- coef(eqc_res)
cat(sprintf("Step 6: Extracted %d item parameters.\n", nrow(item_params)))
```


# 8. Publication-Ready Language

The following text templates can be adapted for the Method section of a journal
paper.  Replace bracketed values with your specific settings.

## 8.1 Describing the simulation design

> Item response data were generated using the reliability-targeted simulation
> framework of Lee (2026), implemented in the R package IRTsimrel (version
> `r packageVersion("IRTsimrel")`). A [Rasch / two-parameter logistic /
> three-parameter logistic] model was assumed with [25] items and latent
> abilities drawn from a [standard normal / bimodal / skewed] distribution.
> For the 3PL, item lower asymptotes were [fixed at 0.20 / generated from the
> stated distribution], and calibration multiplied discrimination only.

## 8.2 Describing the calibration

> The global discrimination scaling factor $c^*$ was calibrated using the
> Empirical Quadrature Calibration algorithm (EQC; Algorithm 1 in Lee, 2026)
> with a Monte Carlo quadrature sample of size $M =$ [10,000] and the
> average-information reliability metric ($\tilde{\rho}$).  The target marginal
> reliability was set to $\rho^* =$ [0.85], and the achieved reliability was
> [0.8500].

## 8.3 Describing optional SAC validation

> EQC results were cross-validated using the Stochastic Approximation
> Calibration algorithm (SAC; Algorithm 2 in Lee, 2026) with [300] Robbins--Monro
> iterations, warm-started from the EQC solution.  The two algorithms agreed to
> within [X]% on the calibrated scaling factor.

## 8.4 Describing the response data

> Binary item response data were generated for $N =$ [500] simulated examinees
> using the calibrated item parameters. Each response $Y_{pi}$ was drawn from
> $\text{Bernoulli}(p_{pi})$ where $p_{pi} = g_i + (1-g_i)
> \text{logit}^{-1}[\lambda_i^*(\theta_p - \beta_i)]$ (with $g_i=0$ for
> Rasch/2PL). Simulation provenance, including metric,
> calibration status, item source, latent shape, and seed, was retained with
> each generated data object.

## 8.5 Citing the package

> Lee, J.-H. (2026). Reliability-Targeted Simulation of Item Response Data:
> Solving the Inverse Design Problem. arXiv:2512.16012v2.
> https://doi.org/10.48550/arXiv.2512.16012


# 9. Further Reading

The IRTsimrel documentation suite includes several companion vignettes:

- `vignette("quick-start")` --- A minimal 5-minute introduction.
- `vignette("latent-distributions")` --- Detailed guide to all 12 latent
  distribution shapes and the pre-standardization principle.
- `vignette("item-parameters")` --- Sources and methods for generating realistic
  item parameters, including the copula method and optional IRW integration.
- `vignette("theory-reliability")` --- Mathematical foundations of $\tilde{\rho}$
  and $\bar{w}$.
- `vignette("algorithm-eqc")` --- Detailed derivation and analysis of Algorithm 1.
- `vignette("algorithm-sac")` --- Detailed derivation and analysis of Algorithm 2.
- `vignette("validation")` --- Comprehensive validation studies comparing
  IRTsimrel outputs against TAM.
- `vignette("api-reference")` --- Complete function reference with all parameters.


# References

Lee, J.-H. (2026). Reliability-Targeted Simulation of Item Response Data:
Solving the Inverse Design Problem. arXiv:2512.16012v2.
https://doi.org/10.48550/arXiv.2512.16012

Robbins, H., & Monro, S. (1951). A stochastic approximation method.
*The Annals of Mathematical Statistics, 22*(3), 400--407.

Polyak, B. T., & Juditsky, A. B. (1992). Acceleration of stochastic
approximation by averaging. *SIAM Journal on Control and Optimization,
30*(4), 838--855.

Baker, F. B., & Kim, S.-H. (2004). *Item Response Theory: Parameter Estimation
Techniques* (2nd ed.). Marcel Dekker.

Sweeney, S. M., et al. (2022). An investigation of the nature and consequence
of the relationship between IRT difficulty and discrimination.
*Educational Measurement: Issues and Practice, 41*(4), 50--67.
