---
title: "Generating Realistic Item Parameters"
author: "JoonHo Lee (jlee296@ua.edu)"
date: "`r Sys.Date()`"
description: >
  A guide to generating realistic item parameters with sim_item_params(),
  covering Rasch, 2PL, and 3PL generation, optional IRW integration,
  hierarchical and custom sources, correlated parameters, lower asymptotes,
  and extracting calibrated items with coef().
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Generating Realistic Item Parameters}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE, comment = "#>",
  fig.width = 7, fig.height = 5,
  fig.align = "center", out.width = "85%"
)
library(IRTsimrel)
set.seed(42)

# Consistent color palette
pal <- list(
  primary    = "steelblue",
  secondary  = "darkorange",
  accent     = "firebrick",
  muted      = "grey60",
  light_fill = adjustcolor("steelblue", alpha.f = 0.3)
)
```

# 1. Overview

**Reading time**: approximately 20--25 minutes.

The `sim_item_params()` function generates item parameters---difficulty
$\beta$, discrimination $\lambda$, and, for the 3PL, lower asymptote
$g$---for IRT simulation studies. It is designed with five key principles:

1. **Realistic difficulties**: Parametric defaults for portable simulations,
   with optional Item Response Warehouse (IRW) integration for empirically
   grounded difficulty distributions when the IRW package is installed.
2. **Correlated parameters**: Support for the empirically observed negative
   correlation between difficulty and discrimination.
3. **Marginal control**: The copula method keeps the sampled difficulty values
   unchanged while approximately targeting the requested discrimination
   marginal and dependence in a finite form.
4. **Reliability targeting**: A scale factor enables subsequent calibration for
   target reliability.
5. **Explicit 3PL guessing**: Lower asymptotes are generated independently of
   the global discrimination scale and retained in downstream results.

This vignette covers:

1. The difficulty-discrimination correlation
2. Basic usage (Rasch, 2PL, and 3PL)
3. Sources for difficulty generation
4. Methods for discrimination generation
5. Customizing discrimination parameters
6. Generating multiple test forms
7. Guessing distributions and the global scale parameter
8. Visualization
9. Extracting calibrated item parameters with `coef()`
10. Integration with calibration functions

For the complete calibration workflow, see `vignette("applied-guide")`.  For
theoretical details on how item parameters interact with reliability, see
`vignette("algorithm-eqc")`.


# 2. The Difficulty-Discrimination Correlation

A critical finding from psychometric research is that item difficulty and
discrimination are **negatively correlated** in real assessments (Sweeney et al.,
2022):

- Easy items ($\beta$ low) tend to have higher discrimination ($\lambda$ high).
- Difficult items ($\beta$ high) tend to have lower discrimination ($\lambda$ low).

This correlation, typically around $\rho \approx -0.3$, has important
implications:

- Ignoring it produces unrealistic simulation data.
- Standard independent generation misses this structural feature.
- The correlation affects test information functions.

`sim_item_params()` handles this by default using the **copula method** with
$\rho = -0.3$.


# 3. Basic Usage

## 3.1 Rasch model

For the Rasch model, all discriminations are set to 1:

```{r rasch-basic}
# Generate 25 Rasch items with parametric difficulties
items_rasch <- sim_item_params(
  n_items = 25,
  model   = "rasch",
  source  = "parametric",
  seed    = 42
)

print(items_rasch)
```

## 3.2 Two-parameter logistic (2PL) model

For the 2PL model, both difficulties and discriminations are generated:

```{r 2pl-basic}
# Generate 30 2PL items with correlated parameters
items_2pl <- sim_item_params(
  n_items = 30,
  model   = "2pl",
  source  = "parametric",
  method  = "copula",
  seed    = 42
)

print(items_2pl)
```

The printed **achieved correlation** is one finite-form realization to compare
with the default latent-Gaussian dependence input of $-0.3$; it is not
constrained to equal that input.

## 3.3 Three-parameter logistic (3PL) model

The 3PL adds an item lower asymptote $g_i$ while retaining the 2PL difficulty
and discrimination generators. Supply its generator through
`guessing_params`; the default is a fixed value of 0.20.

```{r 3pl-basic}
items_3pl <- sim_item_params(
  n_items = 30,
  model = "3pl",
  source = "parametric",
  method = "copula",
  guessing_params = list(distribution = "fixed", value = 0.20),
  seed = 42
)

items_3pl$data[1:6, c("beta", "lambda", "guessing")]
```

Available guessing generators are:

| `distribution` | Fields | Meaning |
|:---------------|:-------|:--------|
| `"fixed"` | `value` | One scalar (recycled) or one value per item |
| `"beta"` | `shape1`, `shape2` | Beta-distributed values on $[0,1)$ |
| `"uniform"` | `min`, `max` | Uniform values with $0 \leq \min < \max < 1$ |

For `source = "custom"`, put the lower asymptotes in
`custom_params$guessing` instead of `guessing_params`. Every $g_i$ must be
finite and satisfy $0 \leq g_i < 1$.

```{r 3pl-g-zero-compatibility}
# With g = 0, the 3PL is exactly nested in the 2PL. The same seed also
# produces the same beta and lambda draws because fixed guessing uses no RNG.
items_2pl_g0 <- sim_item_params(12, model = "2pl", seed = 91)
items_3pl_g0 <- sim_item_params(
  12, model = "3pl",
  guessing_params = list(distribution = "fixed", value = 0),
  seed = 91
)
stopifnot(
  identical(items_2pl_g0$data$beta, items_3pl_g0$data$beta),
  identical(items_2pl_g0$data$lambda, items_3pl_g0$data$lambda),
  all(items_3pl_g0$data$guessing == 0)
)
```


# 4. Sources for Difficulty Generation

`sim_item_params()` supports four sources for generating item difficulties.

## 4.1 Optional IRW (Item Response Warehouse)

IRW is an optional external source for empirically grounded simulations. It can
provide difficulty distributions based on real assessment items when the IRW
package is installed. The default, dependency-free source remains
`source = "parametric"`.

```{r source-irw, eval = FALSE}
if (requireNamespace("irw", quietly = TRUE)) {
  items_irw <- sim_item_params(
    n_items = 25,
    model   = "rasch",
    source  = "irw",
    seed    = 42
  )

  summary(items_irw$data$beta)
}
```

## 4.2 Parametric

Generate difficulties from a parametric distribution:

```{r source-parametric}
# Normal distribution (default)
items_normal <- sim_item_params(
  n_items = 25,
  model   = "rasch",
  source  = "parametric",
  difficulty_params = list(mu = 0, sigma = 1, distribution = "normal"),
  seed    = 42
)

# Uniform distribution
items_uniform <- sim_item_params(
  n_items = 25,
  model   = "rasch",
  source  = "parametric",
  difficulty_params = list(mu = 0, sigma = 1, distribution = "uniform"),
  seed    = 42
)
```

```{r parametric-compare, fig.height = 4, fig.alt = "Side-by-side plots comparing parametric item difficulty and discrimination distributions."}
par(mfrow = c(1, 2))
hist(items_normal$data$beta, breaks = 12, col = pal$primary,
     border = "white", main = "Normal Difficulties", xlab = expression(beta))
hist(items_uniform$data$beta, breaks = 12, col = pal$secondary,
     border = "white", main = "Uniform Difficulties", xlab = expression(beta))
par(mfrow = c(1, 1))
```

## 4.3 Hierarchical

Joint bivariate normal generation following Glas & van der Linden (2003).  Both
$\log(\lambda)$ and $\beta$ are drawn from a multivariate normal:

$$\begin{pmatrix} \log(\lambda_i) \\ \beta_i \end{pmatrix} \sim N\!\left( \begin{pmatrix} \mu_\lambda \\ \mu_\beta \end{pmatrix}, \begin{pmatrix} \tau_\lambda^2 & \rho \tau_\lambda \tau_\beta \\ \rho \tau_\lambda \tau_\beta & \tau_\beta^2 \end{pmatrix} \right)$$

```{r source-hierarchical}
items_hier <- sim_item_params(
  n_items = 25,
  model   = "2pl",
  source  = "hierarchical",
  hierarchical_params = list(
    mu  = c(0, 0),
    tau = c(0.25, 1),
    rho = -0.3
  ),
  seed = 42
)

print(items_hier)
```

## 4.4 Custom

Supply your own parameters directly:

```{r source-custom}
# Custom difficulties and discriminations
items_custom <- sim_item_params(
  n_items = 10,
  model   = "2pl",
  source  = "custom",
  custom_params = list(
    beta   = seq(-2, 2, length.out = 10),
    lambda = rep(1.2, 10)
  ),
  seed = 42
)

items_custom$data
```

A custom 3PL form requires all three item vectors. A scalar guessing value is
accepted and recycled, but an explicit vector is often clearer when freezing a
form for calibration:

```{r source-custom-3pl}
items_custom_3pl <- sim_item_params(
  n_items = 10,
  model = "3pl",
  source = "custom",
  custom_params = list(
    beta = seq(-2, 2, length.out = 10),
    lambda = rep(1.2, 10),
    guessing = rep(0.20, 10)
  ),
  seed = 42
)
items_custom_3pl$data
```

You can also provide functions that generate parameters:

```{r source-custom-function, eval = FALSE}
items_custom_fn <- sim_item_params(
  n_items = 20,
  model   = "2pl",
  source  = "custom",
  custom_params = list(
    beta   = function(n) rnorm(n, 0, 1.5),
    lambda = function(n) rlnorm(n, 0, 0.3)
  ),
  seed = 42
)
```


# 5. Methods for Discrimination Generation

When using `source = "irw"` or `source = "parametric"` with `model = "2pl"`
or `model = "3pl"`,
you can choose how discriminations are generated. The default is `method =
"copula"`, and three methods are available.

## 5.1 Copula method (recommended)

The **rank-based Gaussian copula** method leaves the sampled difficulty values
unchanged while approximately targeting the requested discrimination marginal
and dependence in a finite form. The supplied `rho` is a latent Gaussian
dependence parameter; realized Pearson and Spearman correlations are stochastic
and are not guaranteed to equal it.

**Algorithm:**

1. Transform $\beta$ to uniform scores:
   $u = \{\text{rank}(\beta)-0.5\}/n$.
2. Transform to normal: $z_\beta = \Phi^{-1}(u)$.
3. Generate correlated normal: $z_\lambda = \rho \cdot z_\beta + \sqrt{1-\rho^2} \cdot z_{\text{indep}}$.
4. Transform to uniform: $v = \Phi(z_\lambda)$.
5. Transform to log-normal: $\lambda = \exp(\mu + \sigma \cdot \Phi^{-1}(v))$.

```{r method-copula}
items_copula <- sim_item_params(
  n_items = 100,
  model   = "2pl",
  source  = "parametric",
  method  = "copula",
  discrimination_params = list(
    mu_log    = 0,
    sigma_log = 0.3,
    rho       = -0.3
  ),
  seed = 42
)

# Check achieved correlation
cat(sprintf("Target rho: -0.30\n"))
cat(sprintf("Achieved Spearman: %.3f\n",
            items_copula$achieved$overall$cor_spearman_pooled))
```

**Why copula is recommended:**

- Preserves the exact difficulty distribution.
- Approximately targets a log-normal discrimination marginal in a finite form.
- Uses `rho` as a latent-Gaussian dependence input; realized **Pearson** and
  **Spearman** correlations are stochastic and should be inspected.
- Works well with any difficulty distribution shape.

## 5.2 Conditional method

Uses conditional normal regression:

$$\log(\lambda_i) \mid \beta_i \sim N\!\left(\mu_{\log} + \rho \cdot \sigma_{\log} \cdot z_{\beta_i}, \; \sigma_{\log}\sqrt{1-\rho^2}\right)$$

```{r method-conditional}
items_cond <- sim_item_params(
  n_items = 100,
  model   = "2pl",
  source  = "parametric",
  method  = "conditional",
  discrimination_params = list(rho = -0.3),
  seed    = 42
)

cat(sprintf("Achieved Pearson:  %.3f\n",
            items_cond$achieved$overall$cor_pearson_pooled))
cat(sprintf("Achieved Spearman: %.3f\n",
            items_cond$achieved$overall$cor_spearman_pooled))
```

> **Note:** The conditional method assumes linear relationships and normal
> errors.  When empirical or other non-normal difficulties are used, achieved correlations may
> differ from targets.

## 5.3 Independent method

Generates discriminations independently of difficulties (no correlation):

```{r method-independent}
items_indep <- sim_item_params(
  n_items = 100,
  model   = "2pl",
  source  = "parametric",
  method  = "independent",
  seed    = 42
)

cat(sprintf("Achieved correlation: %.3f (expected: ~0)\n",
            items_indep$achieved$overall$cor_spearman_pooled))
```


# 6. Customizing Discrimination Parameters

The `discrimination_params` list controls the log-normal distribution of
discriminations:

```{r disc-params}
# Higher average discrimination
items_high_disc <- sim_item_params(
  n_items = 30,
  model   = "2pl",
  source  = "parametric",
  method  = "copula",
  discrimination_params = list(
    mu_log    = 0.3,
    sigma_log = 0.25,
    rho       = -0.3
  ),
  seed = 42
)

cat(sprintf("Mean lambda: %.3f\n", mean(items_high_disc$data$lambda)))
cat(sprintf("SD lambda:   %.3f\n", sd(items_high_disc$data$lambda)))
```

## 6.1 Understanding the parameters

| Parameter | Default | Description |
|:----------|:--------|:------------|
| `mu_log` | 0 | Mean of $\log(\lambda)$. $\mathbb{E}[\lambda] \approx \exp(\mu_{\log} + \sigma_{\log}^2/2)$ |
| `sigma_log` | 0.3 | SD of $\log(\lambda)$. Controls heterogeneity across items |
| `rho` | -0.3 | Latent-Gaussian dependence input for $\beta$ and $\log(\lambda)$; finite-form correlations are stochastic |


# 7. Generating Multiple Test Forms

Generate multiple parallel forms with independent item samples:

```{r multiple-forms}
items_5forms <- sim_item_params(
  n_items = 20,
  model   = "2pl",
  source  = "parametric",
  method  = "copula",
  n_forms = 5,
  seed    = 42
)

cat(sprintf("Total items: %d\n", nrow(items_5forms$data)))
cat(sprintf("Items per form: %d\n", items_5forms$n_items))
cat(sprintf("Number of forms: %d\n", items_5forms$n_forms))

# View first few rows
head(items_5forms$data, 10)
```

## 7.1 Per-form statistics

```{r form-stats}
for (f in 1:3) {
  stats <- items_5forms$achieved$by_form[[f]]
  cat(sprintf("Form %d: beta_mean=%.3f, lambda_mean=%.3f, cor=%.3f\n",
              f, stats$beta_mean, stats$lambda_mean, stats$cor_spearman))
}
```


# 8. The Scale Parameter

The `scale` parameter is central to the reliability-targeted framework.  It
multiplies all discriminations by a constant factor:

$$\lambda_i^* = c \cdot \lambda_{i,0}$$

where $\lambda_{i,0}$ is the baseline discrimination and $c$ is the scale
factor.

For a 3PL item, the full response probability is

$$
P(Y_i=1\mid\theta)=g_i+(1-g_i)
\operatorname{logit}^{-1}\{c\lambda_{i,0}(\theta-\beta_i)\}.
$$

Thus, `scale = c` multiplies only discrimination. It does **not** scale the
lower asymptote: use `guessing_params` (or `custom_params$guessing`) to define
$g_i$. Calibration functions estimate this same global $c$ while holding the
generated $\beta_i$ and $g_i$ fixed.

```{r scale-parameter}
# Baseline (scale = 1)
items_base <- sim_item_params(
  n_items = 25, model = "2pl", source = "parametric",
  scale = 1, seed = 42
)

# Scaled up (scale = 1.5)
items_scaled <- sim_item_params(
  n_items = 25, model = "2pl", source = "parametric",
  scale = 1.5, seed = 42
)

cat(sprintf("Baseline mean lambda: %.3f\n", mean(items_base$data$lambda)))
cat(sprintf("Scaled mean lambda:   %.3f\n", mean(items_scaled$data$lambda)))
cat(sprintf("Ratio: %.2f\n",
            mean(items_scaled$data$lambda) / mean(items_base$data$lambda)))
```

## 8.1 Unscaled lambda

The output always includes `lambda_unscaled` for reference:

```{r unscaled-lambda}
head(items_scaled$data[, c("lambda", "lambda_unscaled")])

# Verify relationship
all.equal(
  items_scaled$data$lambda,
  items_scaled$data$lambda_unscaled * items_scaled$scale
)
```

The distinction is directly checkable for a 3PL form:

```{r scale-versus-guessing}
items_3pl_scaled <- sim_item_params(
  n_items = 12,
  model = "3pl",
  scale = 1.5,
  guessing_params = list(distribution = "fixed", value = 0.20),
  seed = 42
)
stopifnot(
  all.equal(
    items_3pl_scaled$data$lambda,
    1.5 * items_3pl_scaled$data$lambda_unscaled
  ),
  all(items_3pl_scaled$data$guessing == 0.20)
)
```


# 9. Centering Difficulties

By default, difficulties are centered to sum to zero (for model identification):

```{r centering}
# Default: centered
items_centered <- sim_item_params(
  n_items = 25, model = "rasch", source = "parametric",
  center_difficulties = TRUE, seed = 42
)

# Uncentered
items_uncentered <- sim_item_params(
  n_items = 25, model = "rasch", source = "parametric",
  center_difficulties = FALSE, seed = 42
)

cat(sprintf("Centered mean:   %.6f\n", mean(items_centered$data$beta)))
cat(sprintf("Uncentered mean: %.6f\n", mean(items_uncentered$data$beta)))
```


# 10. Visualization

The `plot()` method provides diagnostic visualizations:

```{r plot-items, fig.width = 8, fig.height = 6, fig.alt = "Scatter plot of item difficulty and discrimination with diagnostic reference patterns."}
items_viz <- sim_item_params(
  n_items = 50, model = "2pl", source = "parametric",
  method = "copula", seed = 42
)

# Scatter plot with regression line
plot(items_viz, type = "scatter")
```

```{r plot-density, fig.width = 8, fig.height = 4, fig.alt = "Density plots of simulated item difficulty and discrimination parameters."}
# Density plots
plot(items_viz, type = "density")
```


# 11. Extracting Calibrated Item Parameters with `coef()`

After running `eqc_calibrate()` or `sac_calibrate()`, the `coef()` method
extracts a tidy data frame of all item parameters---including both baseline and
calibrated (scaled) discriminations.

## 11.1 From an EQC result

```{r coef-eqc}
eqc_result <- eqc_calibrate(
  target_rho  = 0.85,
  n_items     = 25,
  model       = "rasch",
  item_source = "parametric",
  reliability_metric = "info",
  M           = 5000L,
  seed        = 42
)

# Extract calibrated item parameters
item_table <- coef(eqc_result)
head(item_table)
```

Each row contains:

| Column | Description |
|:-------|:------------|
| `item_id` | Item identifier (1 to $I$) |
| `beta` | Item difficulty |
| `lambda_base` | Baseline (unscaled) discrimination |
| `lambda_scaled` | Calibrated discrimination: $\lambda_{\text{base}} \times c^*$ |
| `c_star` | The calibrated scaling factor (constant across items) |
| `guessing` | 3PL lower asymptote; present for 3PL results and not scaled by $c^*$ |

## 11.2 Summary statistics from coef output

```{r coef-summary}
cat(sprintf("Number of items:       %d\n", nrow(item_table)))
cat(sprintf("Difficulty range:      [%.3f, %.3f]\n",
            min(item_table$beta), max(item_table$beta)))
cat(sprintf("Difficulty mean:       %.4f\n", mean(item_table$beta)))
cat(sprintf("Difficulty SD:         %.4f\n", sd(item_table$beta)))
cat(sprintf("Calibrated c*:         %.4f\n", item_table$c_star[1]))
cat(sprintf("Scaled discrimination: %.4f (all equal for Rasch)\n",
            item_table$lambda_scaled[1]))
```

## 11.3 From a SAC result

The same `coef()` interface works for SAC results:

```{r coef-sac}
sac_result <- sac_calibrate(
  target_rho        = 0.85,
  n_items           = 25,
  model             = "rasch",
  reliability_metric = "info",
  c_init            = eqc_result,
  resample_items    = FALSE,
  n_iter            = 200L,
  M_per_iter        = 500L,
  M_pre             = 5000L,
  seed              = 42
)

sac_items <- coef(sac_result)
head(sac_items)
```

## 11.4 Comparing EQC and SAC item parameters

```{r coef-compare}
cat(sprintf("EQC c*: %.4f\n", item_table$c_star[1]))
cat(sprintf("SAC c*: %.4f\n", sac_items$c_star[1]))
cat(sprintf("Difference: %.4f (%.2f%%)\n",
            abs(item_table$c_star[1] - sac_items$c_star[1]),
            100 * abs(item_table$c_star[1] - sac_items$c_star[1]) / item_table$c_star[1]))
```

## 11.5 Exporting to CSV

```{r coef-export, eval = FALSE}
# Save calibrated item parameters for use in other software
output_file <- file.path(tempdir(), "calibrated_items.csv")
write.csv(coef(eqc_result), file = output_file, row.names = FALSE)
output_file
```

## 11.6 Using coef output with 2PL models

With 2PL models, the baseline discriminations vary across items, and `coef()`
makes it easy to see both the original and scaled values:

```{r coef-2pl}
eqc_2pl <- eqc_calibrate(
  target_rho  = 0.80,
  n_items     = 20,
  model       = "2pl",
  item_source = "parametric",
  reliability_metric = "info",
  M           = 5000L,
  seed        = 42
)

items_2pl_df <- coef(eqc_2pl)
head(items_2pl_df, 10)
```

```{r coef-2pl-plot, fig.height = 4.5, fig.alt = "Scatter plot of calibrated 2PL item difficulties and scaled discriminations."}
plot(items_2pl_df$beta, items_2pl_df$lambda_scaled,
     pch = 16, col = pal$primary, cex = 1.2,
     xlab = expression(beta), ylab = expression(lambda^"*"),
     main = "Calibrated Item Parameters (2PL)")
abline(lm(lambda_scaled ~ beta, data = items_2pl_df),
       col = pal$accent, lty = 2, lwd = 1.5)
```

## 11.7 Using `coef()` with a 3PL model

For 3PL results, the coefficient table retains the fixed lower asymptotes:

```{r coef-3pl}
eqc_3pl <- eqc_calibrate(
  target_rho = 0.70,
  n_items = 12,
  model = "3pl",
  item_params = list(
    guessing_params = list(distribution = "fixed", value = 0.20)
  ),
  reliability_metric = "info",
  M = 2000L,
  seed = 310
)

items_3pl_df <- coef(eqc_3pl)
items_3pl_df[1:6, c(
  "beta", "lambda_base", "lambda_scaled", "guessing", "c_star"
)]
stopifnot(all(items_3pl_df$guessing == 0.20))
```


# 12. Integration with eqc_calibrate

In the reliability-targeted simulation framework, `sim_item_params()` is called
internally by `eqc_calibrate()`.  You specify item generation settings through
the `item_source` and `item_params` arguments:

```{r integration-example}
# EQC automatically calls sim_item_params internally
eqc_result_full <- eqc_calibrate(
  target_rho  = 0.80,
  n_items     = 25,
  model       = "2pl",
  item_source = "parametric",
  reliability_metric = "info",
  item_params = list(
    discrimination_params = list(
      mu_log    = 0,
      sigma_log = 0.3,
      rho       = -0.3
    )
  ),
  M    = 5000L,
  seed = 42
)

cat(sprintf("c* = %.4f, achieved rho = %.4f\n",
            eqc_result_full$c_star, eqc_result_full$achieved_rho))
```

For a 3PL calibration, the guessing generator is nested inside `item_params`
because `eqc_calibrate()` passes that list to `sim_item_params()`:

```{r integration-3pl}
eqc_3pl_generated <- eqc_calibrate(
  target_rho = 0.70,
  n_items = 12,
  model = "3pl",
  item_source = "parametric",
  item_params = list(
    discrimination_params = list(rho = -0.3),
    guessing_params = list(distribution = "beta", shape1 = 5, shape2 = 17)
  ),
  reliability_metric = "info",
  M = 2000L,
  seed = 311
)
summary(coef(eqc_3pl_generated)$guessing)
```

EQC calibrates the one form stored in its result (`item_scope =
"fixed_form"`). SAC can either reuse that form with `resample_items = FALSE`
or target a generator-level item superpopulation with `resample_items = TRUE`.
Those are different estimands; only the fixed-form, same-metric run can use an
EQC result as a comparable warm start.

## 12.1 Accessing calibrated items

After calibration, you can access both baseline and calibrated item parameters:

```{r access-items}
# Baseline items (scale = 1)
items_base_obj <- eqc_result_full$items_base

# Calibrated items (scale = c*)
items_calib_obj <- eqc_result_full$items_calib

# The calibration factor
c_star <- eqc_result_full$c_star

cat(sprintf("Baseline scale:    %d\n", items_base_obj$scale))
cat(sprintf("Calibrated scale:  %.4f\n", items_calib_obj$scale))
```


# 13. Comparison of Methods

Compare the three discrimination generation methods:

```{r compare-methods, fig.width = 9, fig.height = 4, fig.alt = "Comparison plots showing item parameter distributions under different generation methods."}
methods <- c("copula", "conditional", "independent")
results <- list()

for (m in methods) {
  results[[m]] <- sim_item_params(
    n_items = 200, model = "2pl", source = "parametric",
    method = m,
    discrimination_params = list(rho = -0.4),
    seed = 123
  )
}

# Compare achieved correlations
cat("Method Comparison (target rho = -0.4):\n")
cat("======================================\n")
for (m in methods) {
  cat(sprintf("%-12s: Pearson = %+.3f, Spearman = %+.3f\n",
              m,
              results[[m]]$achieved$overall$cor_pearson_pooled,
              results[[m]]$achieved$overall$cor_spearman_pooled))
}
```

The copula method usually tracks the requested direction and magnitude of
dependence more closely than the alternatives in this example, but the
finite-form correlation remains stochastic and is not constrained to equal
`rho`.


# 14. Working with the Output Object

The `item_params` object contains rich information:

```{r output-structure}
items <- sim_item_params(
  n_items = 25, model = "2pl", source = "parametric", seed = 42
)

# Structure
names(items)

# Extract as data frame
df <- as.data.frame(items)
head(df)

# Achieved statistics
items$achieved$overall
```


# 15. Summary Tables

## 15.1 Sources

| Source | Description | Best For |
|:-------|:------------|:---------|
| `parametric` | Normal/uniform difficulties | Controlled experiments and dependency-free examples |
| `irw` | Optional Item Response Warehouse integration | Empirically grounded simulations when IRW is installed |
| `hierarchical` | Joint MVN generation | Bayesian framework |
| `custom` | User-supplied parameters | Specific scenarios |

For 3PL models, every source above can be paired with `guessing_params`, except
that `source = "custom"` uses `custom_params$guessing`.

## 15.2 Methods

| Method | Difficulty Values | Dependence Input | Notes |
|:-------|:------------------|:-----------------|:------|
| `copula` | Preserved exactly | Latent Gaussian `rho` | Approximate discrimination marginal and stochastic realized correlations; **recommended** |
| `conditional` | Preserved exactly | Conditional-regression `rho` | Assumes standardized difficulty enters linearly |
| `independent` | Preserved exactly | None | Realized finite-form correlation fluctuates around zero |


# 16. Practical Recommendations

## 16.1 For realistic simulations

```{r practical-realistic, eval = FALSE}
if (requireNamespace("irw", quietly = TRUE)) {
  items <- sim_item_params(
    n_items = 30, model = "2pl",
    source = "irw",
    method = "copula",
    discrimination_params = list(mu_log = 0, sigma_log = 0.3, rho = -0.3),
    seed = 42
  )
}
```

## 16.2 For controlled experiments

```{r practical-controlled, eval = FALSE}
items <- sim_item_params(
  n_items = 25, model = "2pl",
  source = "parametric",
  difficulty_params = list(mu = 0, sigma = 1),
  method = "conditional",
  discrimination_params = list(mu_log = 0, sigma_log = 0.25, rho = 0),
  seed = 42
)
```

## 16.3 For Bayesian frameworks

```{r practical-bayesian, eval = FALSE}
items <- sim_item_params(
  n_items = 25, model = "2pl",
  source = "hierarchical",
  hierarchical_params = list(mu = c(0, 0), tau = c(0.3, 1), rho = -0.3),
  seed = 42
)
```

## 16.4 For controlled 3PL experiments

```{r practical-3pl, eval = FALSE}
items <- sim_item_params(
  n_items = 25,
  model = "3pl",
  source = "parametric",
  method = "copula",
  discrimination_params = list(mu_log = 0, sigma_log = 0.25, rho = -0.3),
  guessing_params = list(distribution = "fixed", value = 0.20),
  scale = 1,
  seed = 42
)
```

Use `scale = 1` when defining the baseline form, then let EQC or SAC estimate
the global discrimination multiplier. Keep `guessing_params` fixed across
conditions unless lower-asymptote variation is itself part of the design.


# References

Lee, J.-H. (2026). Reliability-Targeted Simulation of Item Response Data:
Solving the Inverse Design Problem. arXiv:2512.16012v2.
https://doi.org/10.48550/arXiv.2512.16012

Glas, C. A. W., & van der Linden, W. J. (2003). Computerized adaptive testing
with item cloning. *Applied Psychological Measurement, 27*(4), 247--261.

Sweeney, S. M., et al. (2022). An investigation of the nature and consequence
of the relationship between IRT difficulty and discrimination.
*Educational Measurement: Issues and Practice, 41*(4), 50--67.

Zhang, L., Liu, Y., Molenaar, D., & Domingue, B. (2025). Realistic Simulation
of Item Difficulties. *PsyArXiv*.
https://osf.io/preprints/psyarxiv/jbhxy_v3/
