---
title: "Quick Start: Your First Calibration in 5 Minutes"
author: "JoonHo Lee (jlee296@ua.edu)"
date: "`r Sys.Date()`"
description: >
  Learn to calibrate item parameters for target reliability, generate response
  data, and extract results in just 5 minutes.
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Quick Start: Your First Calibration in 5 Minutes}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
library(IRTsimrel)
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 5,
  fig.align = "center",
  out.width = "85%"
)
set.seed(42)
```

## 1. Overview

This vignette walks you through a complete IRTsimrel workflow in about five
minutes. By the end you will know how to:

1. **Calibrate** item parameters to hit a target marginal reliability.
2. **Check feasibility** and visualize the reliability curve.
3. **Generate** a simulated binary response dataset.
4. **Extract and inspect** calibrated parameters using S3 methods.
5. **Extend** the same workflow to a 3PL test with lower asymptotes.

**Estimated time:** 5 minutes.

**Prerequisites:** Only the IRTsimrel package is needed. The executable chunks
in the walkthrough run on any system with the current IRTsimrel version
installed; the final full-workflow block is shown as a copyable template.


## 2. Calibrate

The core function is `eqc_calibrate()`. Give it a target reliability, the
number of items, and a measurement model, and it returns a calibrated scaling
factor $c^*$ that makes the selected empirical reliability objective match
your target. Population accuracy remains a Monte Carlo approximation and is
best assessed on independent draws.

```{r calibrate}
result <- eqc_calibrate(
  target_rho = 0.80,
  n_items    = 20,
  model      = "rasch",
  reliability_metric = "info",
  seed       = 42,
  M          = 5000L
)
```

Print the result to see the key quantities:

```{r print-result}
result
```

There are two numbers to focus on in the output:

- **`c*`** (scaling factor): All baseline discriminations are multiplied by
  this value. In a Rasch model the baseline discriminations are all 1, so
  the calibrated discriminations equal $c^*$ directly. A larger $c^*$ means
  the test needs more discriminating items to reach the target.

- **`achieved_rho`**: The empirical reliability at the calibrated $c^*$,
  computed over the Monte Carlo quadrature sample. This should be very close
  to the target of 0.80 --- typically the absolute error is less than
  $10^{-4}$.

The `model` argument accepts `"rasch"` (all discriminations equal), `"2pl"`
(varying discriminations), or `"3pl"` (varying discriminations plus item
lower asymptotes). For this first pass we use the Rasch model so the familiar
v0.2 workflow is unchanged. A compact, executable 3PL version appears in
Section 7.


## 3. Check Feasibility

Before committing to a particular simulation design, it is good practice to
verify that your target reliability is actually achievable. Not all
combinations of test length, model, and latent distribution can produce
every reliability level.

`check_feasibility()` reports the range of achievable reliabilities for a
given configuration:

```{r feasibility}
feas <- check_feasibility(
  n_items   = 20,
  model     = "rasch",
  target_rho = 0.80,
  seed      = 42,
  M         = 5000L,
  verbose   = FALSE
)
feas
```

The output shows two finite-sample ranges, one for each metric. Range
membership is a screening condition, not by itself a calibration guarantee:
inspect the canonical target status, detected roots, admissible-root counts,
and topology resolution. EQC additionally needs a root admitted by its
`root_policy`. SAC performs its own split-bank preflight and requires a
stable interior increasing crossing. It also rejects population-MSEM targeting
for the built-in `heavy_tail` distribution because that functional can be
non-integrable even when a finite grid returns a number.

You can also visualize how reliability varies continuously with the scaling
factor by using `rho_curve()`. This plot shows both the average-information
metric ($\tilde{\rho}$, blue) and the MSEM-based metric ($\bar{w}$, red):

```{r rho-curve, fig.cap="Reliability curve for a 20-item Rasch test. On the same quadrature grid, the average-information metric (blue) lies at or above the MSEM-based metric (red) due to Jensen's inequality.", fig.alt="Reliability curves for a 20-item Rasch test showing the average-information metric above the MSEM-based metric."}
curve_data <- rho_curve(
  n_items = 20,
  model   = "rasch",
  metric  = "both",
  M       = 5000L,
  seed    = 42,
  plot    = TRUE
)
```

The two curves are close together for this configuration. The gap widens for
shorter tests or non-normal latent distributions.


## 4. Generate Data

Once you have a calibration result, `simulate_response_data()` generates a
binary response matrix using the calibrated item parameters. The function
draws fresh latent abilities from the specified distribution and produces
responses according to the stored Rasch, 2PL, or 3PL model with the calibrated
discriminations:

```{r simulate}
sim <- simulate_response_data(
  result    = result,
  n_persons = 500,
  seed      = 123
)
```

The result is a list with five components:

- `response_matrix`: an $N \times I$ matrix of binary (0/1) responses
- `theta`: the true latent abilities for each person
- `beta`: the item difficulties
- `lambda`: the scaled item discriminations
- `guessing`: item lower asymptotes (zero for Rasch/2PL)
- `provenance`: calibration and simulation metadata, including metric, status,
  item source, and seed

```{r dim-check}
# Dimensions: 500 persons x 20 items
dim(sim$response_matrix)
```

```{r provenance-check}
sim$provenance[c("metric", "calibration_status", "status_flags", "item_source")]
```

```{r head-responses}
# First 6 persons, first 8 items
sim$response_matrix[1:6, 1:8]
```

You can verify that the item parameters match what you expect:

```{r check-params}
# All scaled discriminations should equal c*
summary(sim$lambda)

# Difficulty distribution
summary(sim$beta)
```

A quick diagnostic: plot the proportion of correct responses per item. Items
with extreme difficulties should have very high or very low proportions:

```{r item-difficulty-plot, fig.cap="Proportion correct per item, ordered by difficulty. Items near the center of the difficulty distribution have proportions near 0.50, while extreme items show floor or ceiling effects.", fig.alt="Bar plot of item proportion correct ordered by item difficulty."}
p_correct <- colMeans(sim$response_matrix)
item_order <- order(sim$beta)

barplot(
  p_correct[item_order],
  names.arg = item_order,
  col  = "steelblue",
  xlab = "Item (ordered by difficulty)",
  ylab = "Proportion Correct",
  main = "Proportion Correct per Item",
  ylim = c(0, 1),
  las  = 2,
  cex.names = 0.7
)
abline(h = 0.5, lty = 2, col = "gray50")
```


## 5. Extract and Inspect

IRTsimrel result objects support standard S3 generics, so you can interact
with them the same way you would with `lm` or `glm` objects.

### 5.1 Summary

`summary()` returns a compact overview of the calibration:

```{r summary}
summary(result)
```


### 5.2 Extract Coefficients

`coef()` returns a tidy data frame of calibrated item parameters --- one
row per item:

```{r coef}
item_pars <- coef(result)
head(item_pars)
```

For a Rasch model, `lambda_base` is always 1 and `lambda_scaled` equals
$c^*$ for every item. For 2PL and 3PL models, `lambda_base` varies across
items and `lambda_scaled` is `lambda_base * c*`. A 3PL coefficient table also
contains `guessing`; calibration does not multiply that column by $c^*$.


### 5.3 Predict Reliability at New Scaling Factors

`predict()` evaluates the reliability function at new scaling factor values.
With no arguments, it returns the achieved reliability at the calibrated
$c^*$:

```{r predict-default}
# Achieved reliability at c*
predict(result)
```

With `newdata`, it computes reliability at each specified value of $c$:

```{r predict-new}
# Reliability at several scaling factors
predict(result, newdata = c(0.5, 1.0, 1.5, 2.0))
```

This is useful for understanding how sensitive reliability is to the
discrimination level. For instance, at $c = 0.5$ the test is substantially
less reliable, while at $c = 2.0$ it is more reliable than needed.


## 6. Validate (Optional)

For added confidence in your calibration, you can run SAC (Stochastic
Approximation Calibration) as an independent check. SAC uses a completely
different algorithm (Robbins-Monro stochastic approximation). When validating
an EQC result, set `reliability_metric = "info"` so both algorithms solve the
same calibration problem; agreement is then strong evidence that both found
the correct $c^*$.

Passing the EQC result as `c_init` provides a warm start that makes SAC
converge in very few iterations:

```{r sac-validate}
sac_result <- sac_calibrate(
  target_rho = 0.80,
  n_items    = 20,
  model      = "rasch",
  reliability_metric = "info",
  c_init     = result,
  resample_items = FALSE,
  n_iter     = 100L,
  M_per_iter = 500L,
  seed       = 42
)
```

Compare the two algorithms side by side:

```{r compare}
comp <- compare_eqc_sac(result, sac_result)
```

The comparison reports the absolute and percent difference between the two
$c^*$ values and applies its 5% diagnostic agreement threshold. Interpret that
flag together with `comparable`, calibration status, and the independent
holdout reliability rather than as a universal precision guarantee.

You can also visualize the SAC convergence trajectory to verify that the
iterations stabilized near the EQC warm start:

```{r sac-plot, fig.cap="SAC convergence trajectory. The top panel shows the scaling factor c across iterations, and the bottom panel shows the per-iteration reliability estimates. The warm start from EQC ensures rapid convergence.", fig.alt="Two-panel SAC convergence plot showing scaling factor and reliability trajectories across iterations."}
plot(sac_result, type = "both")
```

The top panel shows the scaling factor trajectory, which should stabilize
quickly when initialized from EQC. The bottom panel shows the noisy
per-iteration reliability estimates oscillating around the target value
(dashed red line). The Polyak-Ruppert average (reported as $c^*$) smooths
out the iteration-to-iteration noise.


## 7. Putting It All Together

Here is the complete workflow in a single code block, from calibration to
validated data generation:

```{r full-workflow, eval=FALSE}
library(IRTsimrel)

# Step 1: Check feasibility
feas <- check_feasibility(
  n_items = 20, model = "rasch", target_rho = 0.80,
  seed = 42, verbose = FALSE
)

# Step 2: Calibrate with EQC
eqc_res <- eqc_calibrate(
  target_rho = 0.80, n_items = 20, model = "rasch",
  reliability_metric = "info", seed = 42, M = 5000L
)

# Step 3: Validate with SAC (optional but recommended)
sac_res <- sac_calibrate(
  target_rho = 0.80, n_items = 20, model = "rasch",
  reliability_metric = "info", c_init = eqc_res,
  resample_items = FALSE, n_iter = 100L, seed = 42
)
compare_eqc_sac(eqc_res, sac_res)

# Step 4: Generate response data
sim_data <- simulate_response_data(
  result = eqc_res, n_persons = 1000, seed = 123
)

# Step 5: Use the data in your analysis
dim(sim_data$response_matrix)  # 1000 x 20
```


## 8. The Same Cycle for a 3PL Test

The 3PL response probability is

$$
P(Y_i=1\mid\theta)=g_i+(1-g_i)
\operatorname{logit}^{-1}\{c\lambda_{i,0}(\theta-\beta_i)\}.
$$

Here `guessing_params` generates the lower asymptotes $g_i$. In contrast,
`scale` in `sim_item_params()` and the calibrated $c^*$ are the same kind of
**global discrimination multiplier**: they change $\lambda$, not $g$. The
names are deliberately kept separate in the API to prevent treating a 0.20
guessing parameter as a discrimination scale.

The following small workflow generates a fixed 3PL form, calibrates it with
EQC, validates the same estimand with a fixed-form SAC run, and then generates
responses. The seeds are explicit, and the vignette-sized Monte Carlo settings
keep the example fast.

```{r quick-3pl-workflow}
# Generate and freeze one item form.
items_3pl <- sim_item_params(
  n_items = 12,
  model = "3pl",
  source = "parametric",
  guessing_params = list(distribution = "fixed", value = 0.20),
  seed = 310
)

fixed_3pl <- list(custom_params = list(
  beta = items_3pl$data$beta,
  lambda = items_3pl$data$lambda_unscaled,
  guessing = items_3pl$data$guessing
))

eqc_3pl <- eqc_calibrate(
  target_rho = 0.70,
  n_items = 12,
  model = "3pl",
  item_source = "custom",
  item_params = fixed_3pl,
  reliability_metric = "info",
  M = 2000L,
  seed = 310
)

# Omitted item_source/item_params tell SAC to reuse the EQC form exactly.
sac_3pl <- sac_calibrate(
  target_rho = 0.70,
  n_items = 12,
  model = "3pl",
  reliability_metric = "info",
  c_init = eqc_3pl,
  resample_items = FALSE,
  n_iter = 60L,
  M_per_iter = 250L,
  M_pre = 1000L,
  evaluation_controls = list(n_forms = 3L, M = 500L),
  seed = 310
)

responses_3pl <- simulate_response_data(
  result = eqc_3pl,
  n_persons = 200,
  seed = 311
)

validation_3pl <- compare_eqc_sac(eqc_3pl, sac_3pl)
stopifnot(
  validation_3pl$comparable,
  identical(dim(responses_3pl$response_matrix), c(200L, 12L)),
  all(responses_3pl$guessing == 0.20)
)
head(coef(eqc_3pl))
```

EQC always targets the generated **fixed form**. For SAC,
`resample_items = FALSE` targets that same fixed form, whereas
`resample_items = TRUE` targets an **item superpopulation** by drawing forms
during calibration and reporting the mean reliability across independent
holdout forms. An EQC object is therefore a valid SAC warm start only for the
same metric and fixed item bank; use a numeric or automatic SAC initializer
for a superpopulation run.

Setting all $g_i=0$ gives exact nested 2PL behavior in the response and
information kernels. Existing Rasch/2PL calls need no new arguments, so this
extension is backward compatible.


## 9. What's Next?

You have now completed a full calibrate-generate-validate cycle. Here are
some directions for deeper exploration:

- **`vignette("applied-guide")`**: Comprehensive applied tutorial covering
  Rasch, 2PL, and 3PL models; fixed-form versus item-superpopulation targets;
  non-normal latent distributions; optional IRW-based item sources; and
  factorial simulation designs with multiple reliability levels.

- **`vignette("latent-distributions")`**: Explore all 12 latent distribution
  shapes available in `sim_latentG()` and learn when to use each one for
  different research scenarios.

- **`vignette("item-parameters")`**: Parametric, IRW, hierarchical, and
  custom item generation methods, including correlated difficulty-discrimination
  parameters.

- **`vignette("theory-reliability")`**: Mathematical foundations of the two
  reliability metrics ($\tilde{\rho}$ and $\bar{w}$), Jensen's inequality,
  and the theoretical justification for the calibration approach.

- **`vignette("api-reference")`**: Full function reference with complete
  signatures, all arguments, return values, and runnable examples for every
  exported function.


## References

Lee, J.-H. (2026). Reliability-Targeted Simulation of Item Response Data:
Solving the Inverse Design Problem. arXiv:2512.16012v2.
https://doi.org/10.48550/arXiv.2512.16012
