---
title: "Mathematical Foundations: Reliability Theory and the Inverse Design Problem"
author: "JoonHo Lee (jlee296@ua.edu)"
date: "`r Sys.Date()`"
description: >
  Marginal reliability theory for one-dimensional Rasch, 2PL, and 3PL
  models, including global discrimination scaling, estimand scope, root
  topology, and the contracts implemented by EQC and SAC.
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Mathematical Foundations: Reliability Theory and the Inverse Design Problem}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE, comment = "#>",
  fig.width = 7, fig.height = 5,
  fig.align = "center", out.width = "85%"
)
```

```{r setup}
library(IRTsimrel)
set.seed(42)
```

## Overview

This vignette provides the mathematical foundations for the reliability-targeted
IRT simulation framework implemented in the IRTsimrel package. It develops the
theory from first principles, connecting classical IRT measurement theory to the
inverse calibration problem solved by the EQC and SAC algorithms.

**Reading time**: approximately 45--60 minutes.

**Prerequisites**: Familiarity with basic probability theory (expectation,
variance, convergence), linear algebra, and the logistic function. Some exposure
to item response theory is helpful but not strictly required.

**Notation conventions**: Random variables are denoted by Greek or capital Roman
letters; vectors are boldfaced; hats denote estimators; bars and tildes denote
specific reliability definitions (defined below).

**Cross-references**: For algorithm-specific details, see
`vignette("algorithm-eqc")` and `vignette("algorithm-sac")`. For applied
examples and workflows, see `vignette("introduction")`.


## One-Dimensional Dichotomous Logistic Models

IRTsimrel 0.3.0 works with one-dimensional dichotomous Rasch, 2PL, and 3PL
models. A test consists of $I$ items administered to examinees drawn from a
latent trait distribution $G$.

**Definition 1 (2PL Model).** Let $\theta \sim G$ denote a latent ability
parameter and let $(\beta_i, \lambda_i)$ denote the difficulty and
discrimination parameters for item $i$. The probability of a correct response
is given by the item characteristic curve (ICC):

$$
P(X_i = 1 \mid \theta) = \psi\bigl(\lambda_i(\theta - \beta_i)\bigr)
\qquad \text{(1)}
$$

where $\psi(z) = (1 + e^{-z})^{-1}$ is the standard logistic function.

The Rasch model is the special case where $\lambda_i = 1$ for all items $i$.

**Definition 1b (3PL Model).** Let $g_i\in[0,1)$ be the lower asymptote and
write

$$
q_i(\theta)=\psi\bigl(\lambda_i(\theta-\beta_i)\bigr),\qquad
P_i(\theta)=g_i+(1-g_i)q_i(\theta).
$$

The package calls $g_i$ `guessing`. This name is deliberate: the symbol $c$
is reserved for the global discrimination multiplier introduced below. Setting
$g_i=0$ gives the 2PL probability exactly.

### Item and Test Information

The Fisher information provided by item $i$ at ability level $\theta$ is:

**Definition 2 (Item Information).** For a Bernoulli item with success
probability $P_i(\theta)$,

$$
\mathcal{J}_i(\theta)=
\frac{\{\partial P_i(\theta)/\partial\theta\}^2}
     {P_i(\theta)\{1-P_i(\theta)\}}.
$$

Under the 2PL model this reduces to

$$
\mathcal{J}_i(\theta) = \lambda_i^2\, p_i(\theta)\, \bigl(1 - p_i(\theta)\bigr)
\qquad \text{(2)}
$$

where $p_i(\theta) = P(X_i = 1 \mid \theta)$ is given by Eq. (1).

For the 3PL model it becomes

$$
\mathcal{J}_{i,3\mathrm{PL}}(\theta)
=\lambda_i^2(1-g_i)
  \frac{q_i(\theta)^2\{1-q_i(\theta)\}}{P_i(\theta)}.
$$

The $g_i=0$ reduction is exact. For fixed $\lambda_i,\beta_i$, positive
guessing cannot add Fisher information: it lowers information relative to the
corresponding 2PL item. In the low-ability tail, 2PL information is of order
$\exp(\eta_i)$ whereas 3PL information with $g_i>0$ is of order
$\exp(2\eta_i)$, where $\eta_i=\lambda_i(\theta-\beta_i)$. This faster tail
decay is important for the MSEM estimand.

For Rasch/2PL items, $p(1-p)$ achieves its maximum of $1/4$ when
$p = 1/2$ (i.e., when $\theta = \beta_i$), so the maximum item information is
$\lambda_i^2/4$. This maximum statement does not apply unchanged to $g_i>0$;
the 3PL maximum is shifted and reduced.

**Definition 3 (Test Information).** The test information function is the sum of
item information functions:

$$
\mathcal{J}(\theta) = \sum_{i=1}^{I} \mathcal{J}_i(\theta)
\qquad \text{(3)}
$$

For Rasch/2PL this sum expands to
$\sum_i \lambda_i^2p_i(\theta)\{1-p_i(\theta)\}$; for 3PL it uses the
3PL item-information expression above.

Test information measures the precision of ability estimation at a given
$\theta$. Higher information means smaller standard errors and more precise
measurement.

### The Conditional Standard Error

The asymptotic standard error of the maximum likelihood estimator
$\hat{\theta}$ at $\theta$ is:

$$
\text{SE}(\hat{\theta} \mid \theta) = \frac{1}{\sqrt{\mathcal{J}(\theta)}}
\qquad \text{(4)}
$$

This conditional error varies across the ability continuum, motivating the need
for _marginal_ (population-averaged) reliability measures.


## Two Definitions of Marginal Reliability

Classical test theory defines reliability as the ratio of true-score variance to
observed-score variance. In IRT, multiple marginalizations are possible,
leading to distinct reliability definitions.

Let $\sigma^2_\theta = \text{Var}_G(\theta)$ denote the latent trait variance
under the population distribution $G$.

### Population target and finite approximation

The equations below define population functionals. Package computations use a
finite set of ability nodes (and optional normalized weights) to approximate
those functionals. EQC fixes one Monte Carlo quadrature draw; SAC uses fresh
iteration draws plus an independent post-calibration evaluation. Consequently,
an achieved value is conditional on a finite integration design even when the
scientific target is population reliability. The result metadata records this
distinction through `estimand_signature$theta_measure` and
`estimand_signature$approximation_method`.

If `theta_var` is not supplied to a low-level reliability function, the
unweighted sample variance (or weighted population variance when weights are
provided) is used. Calibration results store the variance actually used. A
generator that is population-standardized does not make every realized sample
have exactly the requested mean and variance.

### Average-Information Reliability (rho-tilde) {#average-information-reliability-tilderho}

**Definition 4 (Average-Information Reliability).** The average-information
marginal reliability is defined as:

$$
\tilde{\rho}(c) = \frac{\sigma^2_\theta\, \bar{\mathcal{J}}(c)}
  {\sigma^2_\theta\, \bar{\mathcal{J}}(c) + 1}
\qquad \text{(5)}
$$

where $\bar{\mathcal{J}}(c) = \mathbb{E}_G[\mathcal{J}(\theta; c)]$ is the
average test information under $G$.

This definition first averages the test information, then converts to a
reliability-like coefficient. It can be interpreted as the reliability one
would obtain if the _average_ measurement precision applied uniformly across all
examinees.

In the package, this metric is selected with `reliability_metric = "info"` (or
the synonym `"tilde"`).
In returned utility objects, this quantity is named `rho_tilde`.

### MSEM-Based Reliability (w-bar) {#msem-based-reliability-barw}

**Definition 5 (MSEM-Based Reliability).** The marginal-standard-error-of-
measurement (MSEM) based reliability is defined as:

$$
\bar{w}(c) = \frac{\sigma^2_\theta}
  {\sigma^2_\theta + \mathbb{E}_G\bigl[1/\mathcal{J}(\theta; c)\bigr]}
\qquad \text{(6)}
$$

This definition first takes the reciprocal of test information at each
$\theta$ (using the asymptotic reciprocal-information approximation to the
conditional error variance), averages these error variances, and then forms the
reliability ratio.

The variance-ratio convention used here is
$\sigma^2_\theta / (\sigma^2_\theta + \text{MSEM})$: the denominator is the
latent true variance plus the population-average reciprocal-information error
variance. Some IRT reliability presentations write related MSEM coefficients as
$1 - \text{MSEM}/\sigma^2_{\text{total}}$; the forms coincide only when the
same total-variance decomposition is used. In this package, Eq. (6) is therefore
the explicit information-based MSEM convention used for reliability-targeted
simulation design, not a claim that all MSEM reliability notation is identical.

In the package, this metric is selected in SAC with
`reliability_metric = "msem"` (or the synonym `"bar"`). EQC targets the
bracket-checked average-information metric instead.
In returned utility objects, this quantity is named `rho_bar`; this API label
denotes the manuscript notation $\bar{w}$, not a separate $\bar{\rho}$
estimand. By contrast, `rho_tilde` corresponds directly to $\tilde{\rho}$.

### Jensen's Inequality: rho-tilde >= w-bar {#jensens-inequality-tilderho-geq-barw}

The relationship between the two metrics is governed by Jensen's inequality.
The comparison assumes that both metrics are evaluated on the same population
distribution, or on the same empirical quadrature grid with the same weights
and the same latent-variance basis, and that
$\mathcal{J}(\theta;c)>0$ with finite $\mathbb{E}[\mathcal{J}]$ and
$\mathbb{E}[1/\mathcal{J}]$.

**Theorem 1 (Jensen's Inequality for Reliability Metrics).** For the population
metrics, and for empirical approximations evaluated on the same finite
quadrature grid, for any scaling factor $c > 0$ and any latent distribution $G$
with $\sigma^2_\theta > 0$,

$$
\tilde{\rho}(c) \geq \bar{w}(c)
\qquad \text{(7)}
$$

with equality if and only if $\mathcal{J}(\theta; c)$ is constant
$G$-almost surely.

**Proof.** Let
$A = \mathbb{E}_G[\mathcal{J}(\theta;c)]$ and
$B = \mathbb{E}_G[1/\mathcal{J}(\theta;c)]$. The function
$f(x) = 1/x$ is strictly convex on $(0, \infty)$, so Jensen's inequality gives
$B \geq 1/A$, with equality if and only if $\mathcal{J}(\theta;c)$ is constant
$G$-almost surely:

$$
B
\geq \frac{1}{A}
\qquad \text{(8)}
$$

Because $x \mapsto \sigma^2_\theta / (\sigma^2_\theta + x)$ is decreasing,

$$
\begin{aligned}
\bar{w}(c)
&= \frac{\sigma^2_\theta}
  {\sigma^2_\theta + B}
\leq \frac{\sigma^2_\theta}
  {\sigma^2_\theta + 1/A} \\
&= \frac{\sigma^2_\theta\, \bar{\mathcal{J}}(c)}
  {\sigma^2_\theta\, \bar{\mathcal{J}}(c) + 1}
= \tilde{\rho}(c)
\end{aligned}
$$

Equality holds iff information does not vary across $\theta$. The
implementation evaluates log test information and uses stable log-sum-exp
reductions; it does not replace small information by an arbitrary positive
floor. Non-finite reciprocal-information expectations are therefore an
estimand failure rather than a finite value manufactured by clipping.
$\square$

### Numerical Verification of Jensen's Inequality

We can verify this inequality computationally using the package's low-level
reliability functions.

```{r jensen-verify}
# Generate a latent trait sample and item parameters
theta <- sim_latentG(5000, shape = "normal")$theta
items <- sim_item_params(25, model = "rasch", source = "parametric")
beta  <- items$data$beta
lambda_base <- rep(1, 25)

# Compute both metrics across a range of c values
c_grid <- seq(0.3, 3.0, by = 0.1)
results <- data.frame(
  c = c_grid,
  rho_tilde = numeric(length(c_grid)),
  rho_bar   = numeric(length(c_grid)),
  gap       = numeric(length(c_grid))
)

for (j in seq_along(c_grid)) {
  both <- compute_rho_both(c_grid[j], theta, beta, lambda_base)
  results$rho_tilde[j] <- both$rho_tilde
  results$rho_bar[j]   <- both$rho_bar
  results$gap[j]       <- both$rho_tilde - both$rho_bar
}

# Verify: gap is non-negative on this common quadrature grid
cat(sprintf("Jensen's gap range: [%.6f, %.6f]\n",
            min(results$gap), max(results$gap)))
cat(sprintf("All gaps >= 0: %s\n", all(results$gap >= -1e-12)))
```

```{r jensen-plot, fig.cap = "Jensen's gap between the two reliability metrics across scaling factors.", fig.alt = "Two-panel plot showing reliability metric curves and their Jensen gap across scaling factors."}
oldpar <- par(mfrow = c(1, 2), mar = c(4.5, 4.5, 3, 1))
on.exit(par(oldpar))

# Panel 1: Both metrics
plot(results$c, results$rho_tilde, type = "l", col = "steelblue", lwd = 2,
     xlab = "Scaling factor c", ylab = "Reliability",
     main = "Two Reliability Metrics", ylim = c(0, 1))
lines(results$c, results$rho_bar, col = "coral", lwd = 2, lty = 2)
legend("bottomright",
       legend = c(expression(tilde(rho) ~ "(info)"),
                  expression(bar(w) ~ "(msem)")),
       col = c("steelblue", "coral"), lty = c(1, 2), lwd = 2, cex = 0.9)

# Panel 2: The gap
plot(results$c, results$gap, type = "l", col = "darkred", lwd = 2,
     xlab = "Scaling factor c", ylab = expression(tilde(rho) - bar(w)),
     main = "Jensen's Gap")
abline(h = 0, lty = 3, col = "gray50")
```

In this example, the gap is largest at intermediate $c$ values where
information varies most across the ability continuum, and shrinks toward zero
for very small $c$ where both metrics approach zero. At very large $c$, the
behavior depends on item-location coverage and the empirical grid: logistic
saturation can concentrate information into narrow neighborhoods of the item
difficulties rather than making information uniformly high.

### Scale Ordering Corollary

**Corollary 1 (Scale Ordering on a Monotone Branch).** Let
$c^*_{\tilde{\rho}}$ and $c^*_{\bar{w}}$ denote calibrated scaling factors for
the same target $\rho^*$ under the average-information and MSEM metrics,
respectively. On a calibration interval where both metrics are continuous and
$\bar{w}(c)$ has a unique strictly increasing root for $\rho^*$, then:

$$
c^*_{\bar{w}} \geq c^*_{\tilde{\rho}}
\qquad \text{(9)}
$$

**Proof sketch.** If $\tilde{\rho}(c^*_{\tilde{\rho}}) = \rho^*$, then
$\bar{w}(c^*_{\tilde{\rho}}) \leq
\tilde{\rho}(c^*_{\tilde{\rho}}) = \rho^*$ by Theorem 1. On a branch where
$\bar{w}(c)$ is continuous and strictly increasing through the target, reaching
$\rho^*$ therefore requires a weakly larger scale. Without this local
monotonicity and unique-root condition, Jensen's inequality gives the pointwise
comparison $\bar{w}(c) \leq \tilde{\rho}(c)$ but does not by itself guarantee a
global ordering of roots, because $\bar{w}(c)$ can be non-monotone. $\square$

This ordering is demonstrated numerically below.

```{r scale-ordering}
# Calibrate the info metric with EQC
eqc_info <- eqc_calibrate(
  target_rho = 0.80, n_items = 25, model = "rasch",
  item_source = "parametric", reliability_metric = "info",
  M = 5000L, seed = 42
)

# Calibrate MSEM on the same realized item form. Because an EQC object carries
# an info-metric contract, pass only its numeric scale as the initializer.
fixed_rasch <- list(
  custom_params = list(
    beta = eqc_info$beta_vec,
    lambda = eqc_info$lambda_base
  ),
  center_difficulties = FALSE
)
sac_msem <- sac_calibrate(
  target_rho = 0.80, n_items = 25, model = "rasch",
  item_source = "custom", item_params = fixed_rasch,
  reliability_metric = "msem", resample_items = FALSE,
  c_init = eqc_info$c_star, n_iter = 200L, M_per_iter = 500L,
  M_pre = 5000L, seed = 42, verbose = FALSE,
  preflight_controls = list(M = 500L),
  evaluation_controls = list(n_forms = 3L, M = 1000L)
)

cat(sprintf("Target rho* = 0.80\n"))
cat(sprintf("  c* (info metric): %.4f\n", eqc_info$c_star))
cat(sprintf("  c* (msem metric): %.4f\n", sac_msem$c_star))
cat(sprintf("  c*_msem >= c*_info: %s\n",
            sac_msem$c_star >= eqc_info$c_star - 1e-6))
```


## Global Discrimination Scaling

The central design variable in IRTsimrel is the global discrimination scaling
factor $c > 0$.

### The Scaling Model

**Definition 6 (Global Discrimination Scaling).** Given baseline item
discriminations $\boldsymbol{\lambda}^{(0)} = (\lambda_1^{(0)}, \ldots,
\lambda_I^{(0)})$ generated from an item parameter distribution $H$, the
scaled discriminations are:

$$
\lambda_i(c) = c \cdot \lambda_i^{(0)}, \quad i = 1, \ldots, I
\qquad \text{(10)}
$$

This scaling has several desirable properties:

1. **Separation of structure and informativeness**: Changing $c$ modifies test
   information without altering the relative difficulty or discrimination
   structure among items.

2. **Dimensionality reduction**: The multi-dimensional item parameter space is
   collapsed to a single scalar control variable $c$.

3. **Smooth dependence on a finite design**: For fixed finite theta nodes and
   a valid item bank, both empirical coefficients are smooth functions of $c$
   for $c > 0$. Extending differentiability through a population expectation
   requires the corresponding domination and integrability conditions; these
   fail for some heavy-tail MSEM settings.

For a 3PL bank, $\boldsymbol{\beta}$ and $\boldsymbol{g}$ remain fixed while
only $\boldsymbol{\lambda}^{(0)}$ is multiplied. Thus the calibration scale
$c$ is not an item guessing parameter. Internally the model is normalized to
the $D=1$ logistic convention.

### Rasch/2PL monotonicity on regular calibration intervals

**Proposition 1 (Monotonicity; cf. Proposition A.1, Lee 2026).** Under the
Rasch/2PL regularity and coverage conditions stated in the manuscript, the
average-information reliability $\tilde{\rho}(c)$ is strictly monotonically
increasing on the population calibration interval.

**Branch-qualified caveat.** This is a population/practical-interval
statement, not a global shape guarantee. A finite empirical quadrature curve
can have several extrema and several target crossings, even for a 2PL bank;
3PL tail attenuation makes global monotonicity still less defensible. Version
0.3.0 therefore scans the complete user-supplied interval on $\log(c)$,
refines detected features, enumerates roots, and selects a branch explicitly.
The default `root_policy = "lowest_increasing"` chooses the smallest detected
crossing with positive local slope. The configured scan has a finite
resolution, so its topology diagnostics are evidence within that resolution,
not a proof that an arbitrarily narrow unobserved feature cannot exist.

**Proof sketch.** A full proof is given in Proposition A.1 of the manuscript.
From Eq. (5), by the chain rule and the fact that
$h(x) = x/(x+1)$ is strictly increasing for $x > 0$, it suffices to establish
that the population average information
$\bar{\mathcal{J}}(c) = \mathbb{E}_G[\mathcal{J}(\theta; c)]$ increases on the
regular target interval.

For a Rasch/2PL item, direct differentiation gives

$$
\frac{\partial \mathcal{J}_i(\theta; c)}{\partial c}
= 2c\, (\lambda_i^{(0)})^2\, p_i(1-p_i)
  + c^2\, (\lambda_i^{(0)})^2\, p_i(1-p_i)(1-2p_i) \cdot \lambda_i^{(0)}(\theta - \beta_i)
$$

The sign of the pointwise derivative need not be positive at every
$\theta$; the population result is an integrated statement. The manuscript's
argument uses the representation
$\bar{\mathcal{J}}(c) = c^2 \sum_i (\lambda_i^{(0)})^2
\mathbb{E}[p_i(\theta;c)(1-p_i(\theta;c))]$
with fixed item locations, positive baseline discriminations, and regular
population mass around the item locations. Under these conditions, increasing
$c$ raises the population-average information over the stated calibration
interval, and therefore increases $\tilde{\rho}(c)$. Outside that interval the
package relies on the observed root topology rather than extending the theorem
globally. $\square$

This proof representation is not the derivative of the 3PL information
function when $g_i>0$. Version 0.3.0 makes no corresponding global 3PL
monotonicity claim; 3PL calibration uses the same numerical topology scan and
requires the selected root's local direction to be admissible.

### Existence and Uniqueness

**Corollary 2 (Existence and Uniqueness; cf. Corollary 2.1, Lee 2026).**
Under the same regularity conditions, consider a practical calibration interval
$[c_L, c_U]$ on which $\tilde{\rho}(c)$ is continuous and strictly increasing.
For any target $\rho^* \in (\rho_L, \rho_U)$ where

$$
\rho_L = \tilde{\rho}(c_L)
\quad\text{and}\quad
\rho_U = \tilde{\rho}(c_U),
$$

there exists a unique $c^* \in (c_L, c_U)$ satisfying
$\tilde{\rho}(c^*) = \rho^*$.

**Proof.** Since $\tilde{\rho}$ is continuous (composition of continuous
functions) and strictly increasing on the interval (Proposition 1), existence
and uniqueness follow from the intermediate value theorem and strict
monotonicity. In the idealized population limit with adequate difficulty
coverage, expanding the interval can approach the familiar range from near-zero
to near-one reliability; the package nevertheless reports feasibility for the
user-supplied finite interval. $\square$

### Non-Monotonicity of MSEM Reliability {#non-monotonicity-of-barwc}

**Proposition 2 (Non-Monotonicity of MSEM; cf. Proposition A.5, Lee 2026).**
The MSEM-based reliability $\bar{w}(c)$ is *not* necessarily monotone in $c$
over the entire positive real line. In particular, for sufficiently large $c$,
$\bar{w}(c)$ can decrease.

**Intuition.** Under Rasch/2PL, as $c \to \infty$, for most $\theta$ values far
from any $\beta_i$, the item response probabilities $p_i(\theta)$ approach 0
or 1, causing item information
$\mathcal{J}_i(\theta)=\lambda_i^2p_i(1-p_i)\to0$. Under 3PL the low tail
probability approaches $g_i$, but its derivative and information still vanish;
positive guessing makes that low-tail information attenuation faster.
Only $\theta$ values very close to some $\beta_i$ retain high information. The
harmonic mean (used in $\bar{w}$) is dominated by the low-information regions,
causing $\mathbb{E}[1/\mathcal{J}]$ to grow and $\bar{w}$ to decrease.

The arithmetic mean (used in $\tilde{\rho}$) is less sensitive to these low-
information tails, but that fact alone does not imply global monotonicity on a
finite grid.

This is why the `c_bounds` parameter matters in practice, especially when using
the `"msem"` metric.

### Tail integrability is part of the estimand

For any finite logistic item bank, $1/\mathcal{J}(\theta;c)$ grows at least
exponentially in a sufficiently remote tail. A Student-$t$ population has only
polynomial tail decay, so
$\mathbb{E}_G[1/\mathcal{J}(\theta;c)]=\infty$ for every $c>0$. Positive 3PL
guessing makes the low-ability divergence faster. It follows that population
MSEM reliability is zero/undefined for the built-in `heavy_tail` population;
a finite Monte Carlo prefix is an empirical truncated approximation and must
not be reported as that population estimand. `sac_calibrate()` rejects the
combination `latent_shape = "heavy_tail"` and
`reliability_metric = "msem"` before iteration. Average-information
reliability remains a separate, potentially finite target.

```{r non-monotone-msem, fig.cap = "Non-monotonicity of the MSEM metric at extreme scaling factors (heavy-tailed latent distribution, df = 3).", fig.alt = "Line plot showing MSEM reliability becoming non-monotone at extreme scaling values under a heavy-tailed latent distribution."}
# Demonstrate non-monotonicity with a heavy-tailed distribution.
# Non-monotonicity is most prominent with heavy tails because extreme
# theta values far from any beta_i have near-zero information, and
# the harmonic mean (used in rho_bar) is dominated by these regions.
theta_nm <- sim_latentG(5000, shape = "heavy_tail", shape_params = list(df = 3))$theta

# Use 10 items with concentrated difficulties for a more visible effect
items_nm <- sim_item_params(10, model = "rasch", source = "parametric")
beta_nm  <- items_nm$data$beta
lambda_nm <- rep(1, 10)

c_wide <- seq(0.1, 15, length.out = 100)
rho_t <- rho_b <- numeric(length(c_wide))

for (j in seq_along(c_wide)) {
  both <- compute_rho_both(c_wide[j], theta_nm, beta_nm, lambda_nm)
  rho_t[j] <- both$rho_tilde
  rho_b[j] <- both$rho_bar
}

oldpar <- par(mar = c(4.5, 4.5, 3, 1))
on.exit(par(oldpar))

plot(c_wide, rho_t, type = "l", col = "steelblue", lwd = 2,
     xlab = "Scaling factor c", ylab = "Reliability",
     main = "Monotonicity Comparison (Heavy-Tailed Latent, df = 3)",
     ylim = c(0, 1))
lines(c_wide, rho_b, col = "coral", lwd = 2, lty = 2)

# Mark the maximum of rho_bar
idx_max <- which.max(rho_b)
points(c_wide[idx_max], rho_b[idx_max], pch = 19, col = "coral", cex = 1.5)

legend("right",
       legend = c(expression(tilde(rho) ~ "(empirical info curve)"),
                  expression(bar(w) ~ "(non-monotone at extreme c)")),
       col = c("steelblue", "coral"), lty = c(1, 2), lwd = 2, cex = 0.85)

cat(sprintf("rho_bar maximum at c = %.2f with value %.4f\n",
            c_wide[idx_max], rho_b[idx_max]))
cat(sprintf("rho_bar at c = %.1f: %.4f (decreasing)\n",
            max(c_wide), rho_b[length(c_wide)]))
```


## Achievable Reliability Bounds

Not every target reliability is achievable for a given test configuration.
The feasible range depends on the number of items, the item parameter
distribution, and the latent trait distribution.

### A 2PL Upper Bound from the Logistic Kernel

**Proposition 3 (Information Upper Bound).** Under the 2PL model, the maximum
item information is:

$$
\sup_{\theta}\, \mathcal{J}_i(\theta) = \frac{\lambda_i^2}{4}
\qquad \text{(11)}
$$

achieved when $\theta = \beta_i$ (i.e., $p_i = 1/2$). Therefore, test
information satisfies:

$$
\sup_{\theta}\, \mathcal{J}(\theta; c)
\;\leq\; \frac{c^2}{4} \sum_{i=1}^{I} (\lambda_i^{(0)})^2
\qquad \text{(12)}
$$

Equality in (12) requires the itemwise maxima to occur at the same ability
value---in particular, all relevant difficulties must coincide. For the Rasch
model with $\lambda_i^{(0)} = 1$, the right-hand upper bound simplifies to
$c^2 I / 4$; it is not generally the attained maximum when difficulties vary.

This $1/4$ bound is a Rasch/2PL result. With 3PL guessing, the information
maximum is shifted and reduced, so IRTsimrel evaluates the realized 3PL bank
directly instead of substituting the 2PL bound.

**Proof.** The function $p(1-p)$ on $[0,1]$ has a unique maximum of $1/4$ at
$p = 1/2$. Under the 2PL model, $p_i(\theta) = 1/2$ when
$\theta = \beta_i$. Summing the itemwise bounds gives (12), while simultaneous
equality requires a common maximizing $\theta$. $\square$

### A 2PL analytic feasibility bound

For Rasch/2PL, the average test information is bounded above by the maximum:

$$
\bar{\mathcal{J}}(c) = \mathbb{E}_G\!\left[\sum_{i=1}^{I} \lambda_i(c)^2\, p_i(1-p_i)\right]
\leq \frac{c^2}{4} \sum_{i=1}^{I} (\lambda_i^{(0)})^2
$$

Therefore an analytic upper bound on average-information reliability for a
given $c$ is:

$$
\tilde{\rho}_{\max}(c) = \frac{\sigma^2_\theta \cdot (c^2/4) \sum_i (\lambda_i^{(0)})^2}
  {\sigma^2_\theta \cdot (c^2/4) \sum_i (\lambda_i^{(0)})^2 + 1}
$$

This upper bound approaches 1 as $c\to\infty$; that does not imply the actual
finite-form reliability curve is monotone or attains 1. For 3PL and for actual
feasibility decisions, the package evaluates the realized curve on the
configured interval.

### Connection to `check_feasibility()`

The `check_feasibility()` function uses a shared adaptive log-scale topology
scan. The achievable range combines boundaries with polished detected extrema;
the result includes root counts, target status, best-achievable points, and the
scanner diagnostics. Endpoint values alone are retained as legacy diagnostics
and are not assumed to determine the global range.

```{r feasibility}
# Check feasibility for a 25-item Rasch test
feas <- check_feasibility(
  n_items = 25, model = "rasch",
  latent_shape = "normal",
  item_source = "parametric",
  target_rho = 0.80,
  c_bounds = c(0.1, 10),
  M = 5000L, seed = 42
)
```

```{r feasibility-targets}
# Test which targets are feasible
targets <- c(0.50, 0.70, 0.80, 0.90, 0.95)
for (rho in targets) {
  feas_rho <- check_feasibility(
    n_items = 25, model = "rasch",
    latent_shape = "normal",
    item_source = "parametric",
    target_rho = rho,
    c_bounds = c(0.1, 10),
    M = 5000L, seed = 42,
    verbose = FALSE
  )
  cat(sprintf("  rho* = %.2f: info %s, msem %s\n",
              rho,
              feas_rho$target_status_info,
              feas_rho$target_status_msem))
}
```

### Reliability Curves

The `rho_curve()` function visualizes how reliability changes with the
scaling factor.

```{r rho-curve, fig.cap = "Reliability curve for a 25-item Rasch test under a normal latent distribution.", fig.alt = "Reliability curve for a 25-item Rasch test under a normal latent distribution."}
curve_data <- rho_curve(
  c_values = seq(0.1, 5, length.out = 60),
  n_items = 25, model = "rasch",
  latent_shape = "normal",
  item_source = "parametric",
  metric = "both",
  M = 5000L, seed = 42,
  plot = TRUE
)
```


## EQC Theory: Deterministic Root-Finding

The Empirical Quadrature Calibration (EQC) algorithm solves the inverse
reliability problem using deterministic root-finding on an empirical
reliability function. Full algorithmic details are in
`vignette("algorithm-eqc")`; here we state the main theoretical results.

### Algorithm Sketch

EQC proceeds in three steps:

1. **Quadrature sampling**: Draw $\{\theta_m\}_{m=1}^M \sim G$ and item
   parameters $\{(\beta_i, \lambda_i^{(0)})\}_{i=1}^I \sim H$ once.

2. **Empirical reliability function**: For any $c > 0$, define:
   $$
   \hat{\tilde{\rho}}_M(c) = \tilde{\rho}\bigl(c;\, \{\theta_m\},\, \boldsymbol{\beta},\,
   \boldsymbol{\lambda}(c)\bigr)
   \qquad \text{(13)}
   $$

3. **Topology and root selection**: Evaluate an adaptive grid on
   $\log(c)$, refine detected extrema and target crossings, polish each
   sign-changing bracket with Brent's method, and select an admissible root
   according to `root_policy`. The default is the lowest increasing crossing.

The scanner never expands `c_bounds`. Its root table distinguishes crossings,
tangencies, boundaries, and plateaus and records local slope direction. If a
target is outside the detected range, EQC returns the best-achievable point
with an explicit infeasibility status; if topology is unresolved or a feasible
target has no root admissible under the requested policy, it raises a classed
error rather than silently choosing another branch.

### Consistency

**Theorem 2 (Branch-conditional consistency of EQC; cf. Theorem A.1, Lee
2026).** Let $c^*$ be an isolated population root on a specified increasing
branch, and let $\hat{c}^*_M$ be the corresponding EQC root with quadrature
size $M$. Under uniform convergence and local root-separation conditions:

$$
\hat{c}^*_M \xrightarrow{\text{a.s.}} c^* \quad \text{as } M \to \infty
\qquad \text{(14)}
$$

**Proof sketch.** The empirical reliability $\hat{\tilde{\rho}}_M(c)$
converges uniformly to $\tilde{\rho}(c)$ over the compact interval
$[c_{\min}, c_{\max}]$ by the Uniform Law of Large Numbers (ULLN). Since
$\tilde{\rho}(c)$ is continuous and strictly monotone on that population
branch, and the selected root is isolated, uniform convergence of the objective
implies convergence of the matching empirical root. This is not a theorem that
the whole configured interval has a unique root. $\square$

### Asymptotic Normality

**Theorem 3 (Asymptotic Normality of EQC; cf. Theorem A.2, Lee 2026).** Under
regularity conditions, the EQC estimator satisfies:

$$
\sqrt{M}\,\bigl(\hat{c}^*_M - c^*\bigr) \xrightarrow{d}
N\!\left(0,\, \frac{\sigma^2_g}{g'(c^*)^2}\right)
\qquad \text{(15)}
$$

where $g(c) = \rho(c) - \rho^*$, $g'(c^*)$ is the derivative of the
reliability function at the solution, and $\sigma^2_g$ is the asymptotic
variance of the empirical estimating-function value
$\hat{\rho}_M(c^*)-\rho^*$ (the root residual), not a variance of the
derivative $g'(c^*)$.

**Proof sketch.** Apply the delta method to the implicit function
$\hat{c}^*_M = g^{-1}_M(\rho^*)$. The Central Limit Theorem gives
$\sqrt{M}(\hat{\rho}_M(c) - \rho(c)) \to_d N(0, \sigma^2_g)$, and the
implicit function theorem (applicable since $g'(c^*) \neq 0$ by strict
monotonicity) yields the result. $\square$

### Monte Carlo Error Bounds

The practical implication of Theorem 3 is:

$$
|\hat{c}^*_M - c^*| = O_p\!\left(\frac{1}{\sqrt{M}}\right)
\qquad \text{(16)}
$$

The reliability estimate itself has error:

$$
|\hat{\rho}_M(c) - \rho(c)| = O_p\!\left(\frac{1}{\sqrt{M}}\right)
$$

The constants depend on the latent distribution, item bank, target, and local
slope; $1/\sqrt{M}$ is a rate statement, not a universal numerical error table.
Independent holdout evaluation is needed because evaluating the selected root
on the same quadrature grid partly verifies the equation by construction.


## SAC Theory: Stochastic Approximation

The Stochastic Approximation Calibration (SAC) algorithm provides an
independent solution path using the Robbins-Monro framework. Full algorithmic
details are in `vignette("algorithm-sac")`.

SAC has two distinct item measures. With `resample_items = FALSE`, it targets
reliability conditional on one fixed item form. With
`resample_items = TRUE`, it targets the mean of form-level reliabilities over
the specified item generator; pooling information across forms would define a
different estimand and is not used. The final `achieved_rho` comes from an
independent evaluation bank. `compare_eqc_sac()` reports agreement only when
metric, latent measure, model, item scope, and realized/generator design are
comparable; otherwise `agreement` is `NA` with reasons.

### The Stochastic Root-Finding Problem

SAC targets the equation:

$$
g(c) = \mathbb{E}\!\left[\hat{\rho}(c)\right] - \rho^* = 0
\qquad \text{(17)}
$$

At each iteration $n$, only a noisy observation
$\hat{\rho}_n(c_n) = \rho^* + g(c_n) + \varepsilon_n$ is available, where
$\varepsilon_n$ is zero-mean Monte Carlo noise.

Before iteration, version 0.3.0 uses common random numbers to scan the
configured interval and identify a stable interior increasing branch. For an
item-superpopulation target, split banks must select compatible roots and
overlapping branch intervals. The Robbins--Monro projection is then narrowed
to that branch. Non-finite topology, an infeasible target, a decreasing or
boundary-only solution, or unstable splits produce classed preflight errors.
Repeated projection hits can instead return a `branch_lost` non-convergence
status. These guards are part of the implemented algorithm, not merely plotting
diagnostics.

### Robbins-Monro Update

The iterative update rule is:

$$
c_{n+1} = \Pi_{[c_L, c_U]}\!\left[c_n - a_n\,\bigl(\hat{\rho}_n(c_n) - \rho^*\bigr)\right]
\qquad \text{(18)}
$$

where $\Pi_{[c_L, c_U]}$ denotes projection onto the constraint set and the
step size sequence satisfies the Robbins-Monro conditions:

$$
a_n = \frac{a}{(n + A)^\gamma}, \quad
\sum_{n=1}^{\infty} a_n = \infty, \quad
\sum_{n=1}^{\infty} a_n^2 < \infty
\qquad \text{(19)}
$$

These conditions are satisfied when $\gamma \in (1/2, 1]$.

### Polyak-Ruppert Averaging

Rather than using the final iterate $c_N$, SAC computes the Polyak-Ruppert
average over post-burn-in iterates:

$$
\bar{c}_N = \frac{1}{N - B} \sum_{n=B+1}^{N} c_n
\qquad \text{(20)}
$$

where $B$ is the burn-in period. Under the standard regular
stochastic-approximation conditions used below, this averaging attains the
usual root-$n$ rate.

### Convergence

**Theorem 4 (Branch-conditional SAC Convergence; cf. Theorem A.3, Lee
2026).** Suppose the Robbins--Monro step conditions (Eq. 19) hold; $c^*$ is
an interior isolated root of the projected selected branch; $g$ is Lipschitz
there and satisfies the branch-wide stability condition
$(c-c^*)g(c)>0$ for every $c\ne c^*$ in that projection interval; and the
Monte Carlo error is a martingale difference with bounded conditional
variance. Then the projected SAC iterates converge almost surely:

$$
c_n \xrightarrow{\text{a.s.}} c^* \quad \text{as } n \to \infty
\qquad \text{(21)}
$$

Furthermore, the Polyak-Ruppert average satisfies:

$$
\sqrt{N - B}\,(\bar{c}_N - c^*) \xrightarrow{d}
N\!\left(0,\, \frac{\sigma^2_\varepsilon}{g'(c^*)^2}\right)
\qquad \text{(22)}
$$

giving root error $O_p(n^{-1/2})$ under the additional CLT regularity
conditions, compared with the typical $O_p(n^{-\gamma/2})$ last-iterate error
for a step sequence of order $n^{-\gamma}$.

**Proof sketch.** Almost sure convergence follows from the classical
Robbins-Monro theorem (Robbins and Monro, 1951), extended to the projected
case by Kushner and Yin (2003). The key conditions are: (i) $g(c^*) = 0$,
(ii) the branch-wide stability condition above together with
$g'(c^*) > 0$ locally, (iii) the step-size conditions in Eq. (19), and
(iv) bounded conditional variance
$\text{Var}(\varepsilon_n \mid c_n) \leq \sigma^2_\varepsilon$. Asymptotic
normality of the Polyak-Ruppert average follows from Polyak and Juditsky
(1992). The theorem is conditional on the selected branch and does not rescue
a run whose branch is absent, non-integrable, or lost. A finite configured
preflight scan is diagnostic evidence, not a proof of the branch-wide drift
assumption. $\square$


## Numerical Demonstrations

### Verifying Theoretical Predictions

We use the low-level reliability functions to verify key theoretical
properties.

```{r demo-setup}
# Common setup for demonstrations
M_demo <- 5000L
theta_demo <- sim_latentG(M_demo, shape = "normal")$theta
items_demo <- sim_item_params(30, model = "rasch", source = "parametric")
beta_demo  <- items_demo$data$beta
lambda_demo <- rep(1, 30)
theta_var_demo <- var(theta_demo)
```

#### Verification 1: Reliability increases with c (info metric)

```{r verify-monotone}
c_test <- c(0.5, 1.0, 1.5, 2.0, 2.5, 3.0)
rho_info <- sapply(c_test, function(cc) {
  compute_rho_tilde(cc, theta_demo, beta_demo, lambda_demo,
                    theta_var = theta_var_demo)
})

cat("Monotonicity check (info metric):\n")
for (j in seq_along(c_test)) {
  cat(sprintf("  c = %.1f: rho_tilde = %.4f", c_test[j], rho_info[j]))
  if (j > 1) {
    cat(sprintf("  (diff = %+.4f)", rho_info[j] - rho_info[j-1]))
  }
  cat("\n")
}
cat(sprintf("All differences positive: %s\n", all(diff(rho_info) > 0)))
```

#### Verification 2: Jensen's gap depends on information heterogeneity

Non-standard latent distributions produce larger Jensen's gaps because
information varies more across the ability continuum.

```{r verify-gap-shapes}
shapes <- c("normal", "bimodal", "heavy_tail", "skew_pos")
shape_pars <- list(
  normal     = list(),
  bimodal    = list(delta = 0.9),
  heavy_tail = list(df = 5),
  skew_pos   = list(k = 4)
)

c_val <- 1.5
cat(sprintf("Jensen's gap at c = %.1f for different latent shapes:\n", c_val))

for (sh in shapes) {
  theta_sh <- sim_latentG(M_demo, shape = sh, shape_params = shape_pars[[sh]])$theta
  both_sh <- compute_rho_both(c_val, theta_sh, beta_demo, lambda_demo)
  gap <- both_sh$rho_tilde - both_sh$rho_bar
  cat(sprintf("  %-12s: rho_tilde = %.4f, rho_bar = %.4f, gap = %.4f\n",
              sh, both_sh$rho_tilde, both_sh$rho_bar, gap))
}
```

#### Verification 3: APC initialization quality

The Analytic Pre-Calibration (APC) formula provides a closed-form starting
point. We compare it against the EQC solution.

```{r verify-apc}
targets <- c(0.60, 0.70, 0.80, 0.85, 0.90)
cat("APC initialization vs EQC solution (Rasch, 25 items, normal):\n")
cat(sprintf("  %-10s %-10s %-10s %-10s\n", "Target", "c_APC", "c_EQC", "Ratio"))

for (rho_t in targets) {
  c_apc <- compute_apc_init(target_rho = rho_t, n_items = 25)
  eqc_t <- eqc_calibrate(
    target_rho = rho_t, n_items = 25, model = "rasch",
    item_source = "parametric", M = 5000L, seed = 42, verbose = FALSE
  )
  cat(sprintf("  %-10.2f %-10.4f %-10.4f %-10.2f\n",
              rho_t, c_apc, eqc_t$c_star, c_apc / eqc_t$c_star))
}
```

#### Verification 4: Effect of test length on achievable reliability

```{r verify-test-length}
n_items_grid <- c(10, 15, 20, 25, 30, 40, 50)
c_fixed <- 1.0

cat(sprintf("Reliability at c = %.1f for different test lengths:\n", c_fixed))
for (ni in n_items_grid) {
  items_ni <- sim_item_params(ni, model = "rasch", source = "parametric")
  beta_ni  <- items_ni$data$beta
  lambda_ni <- rep(1, ni)

  both_ni <- compute_rho_both(c_fixed, theta_demo, beta_ni, lambda_ni,
                               theta_var = theta_var_demo)
  cat(sprintf("  I = %2d: rho_tilde = %.4f, rho_bar = %.4f\n",
              ni, both_ni$rho_tilde, both_ni$rho_bar))
}
```

### EQC vs SAC Agreement

```{r eqc-sac-agree}
eqc_demo <- eqc_calibrate(
  target_rho = 0.80, n_items = 25, model = "rasch",
  item_source = "parametric", reliability_metric = "info",
  M = 5000L, seed = 42, verbose = FALSE
)

sac_demo <- sac_calibrate(
  target_rho = 0.80, n_items = 25, model = "rasch",
  reliability_metric = "info", resample_items = FALSE,
  c_init = eqc_demo, n_iter = 300L, M_per_iter = 1000L,
  seed = 42, verbose = FALSE,
  preflight_controls = list(M = 500L),
  evaluation_controls = list(n_forms = 3L, M = 1000L)
)

cat("EQC vs SAC comparison (target = 0.80, info metric):\n")
cat(sprintf("  EQC c*:        %.4f\n", eqc_demo$c_star))
cat(sprintf("  SAC c*:        %.4f\n", sac_demo$c_star))
cat(sprintf("  Abs diff:      %.4f\n", abs(eqc_demo$c_star - sac_demo$c_star)))
cat(sprintf("  Pct diff:      %.2f%%\n",
            100 * abs(eqc_demo$c_star - sac_demo$c_star) / eqc_demo$c_star))
```

### Verification 5: Reliability function under the 2PL model

The 2PL model introduces additional variability through heterogeneous
baseline discriminations.

```{r verify-2pl}
items_2pl <- sim_item_params(25, model = "2pl", source = "parametric")
beta_2pl  <- items_2pl$data$beta
lambda_2pl <- if ("lambda_unscaled" %in% names(items_2pl$data)) {
  items_2pl$data$lambda_unscaled
} else {
  items_2pl$data$lambda
}

cat("2PL baseline discriminations:\n")
cat(sprintf("  Range: [%.3f, %.3f]\n", min(lambda_2pl), max(lambda_2pl)))
cat(sprintf("  Mean:  %.3f\n", mean(lambda_2pl)))
cat(sprintf("  SD:    %.3f\n", sd(lambda_2pl)))

# Compare Rasch and 2PL reliability at same c
c_compare <- 1.0
both_rasch <- compute_rho_both(c_compare, theta_demo,
                                beta_demo[1:25], rep(1, 25),
                                theta_var = theta_var_demo)
both_2pl <- compute_rho_both(c_compare, theta_demo,
                              beta_2pl, lambda_2pl,
                              theta_var = theta_var_demo)

cat(sprintf("\nReliability comparison at c = %.1f:\n", c_compare))
cat(sprintf("  Rasch: rho_tilde = %.4f, rho_bar = %.4f, gap = %.4f\n",
            both_rasch$rho_tilde, both_rasch$rho_bar,
            both_rasch$rho_tilde - both_rasch$rho_bar))
cat(sprintf("  2PL:   rho_tilde = %.4f, rho_bar = %.4f, gap = %.4f\n",
            both_2pl$rho_tilde, both_2pl$rho_bar,
            both_2pl$rho_tilde - both_2pl$rho_bar))
```

Heterogeneous discriminations typically produce a larger Jensen's gap because
the test information varies more across the ability continuum.


## The Inverse Design Problem

The theoretical framework above describes the _forward_ problem: given a
scaling factor $c$, compute the reliability $\rho(c)$. The central
contribution of the IRTsimrel package is solving the _inverse_ problem.

### Problem Statement

**Definition 7 (Inverse Reliability Problem).** Given a target marginal
reliability $\rho^* \in (0, 1)$, find a global discrimination scaling factor
$c^* > 0$ such that:

$$
\rho(c^*) = \rho^*
\qquad \text{(23)}
$$

where $\rho$ denotes either $\tilde{\rho}$ (average-information) or $\bar{w}$
(MSEM-based), depending on the chosen metric. In IRTsimrel, EQC solves the
branch-qualified $\tilde{\rho}$ inverse problem, while SAC can solve either
same-estimand validation problems (`"info"`) or direct MSEM problems
(`"msem"`).

### Why the Inverse Problem is Difficult

Several factors make this inverse problem non-trivial:

1. **No closed-form solution**: The reliability function involves a logistic
   nonlinearity inside an expectation over $G$. Even in the simplest Rasch
   case with normal $G$, no analytic inversion is possible.

2. **Distribution dependence**: The mapping $c \mapsto \rho(c)$ depends on
   the full distributions $G$ (latent traits) and $H$ (item parameters), not
   just their moments.

3. **Metric ambiguity**: The two reliability definitions yield different $c^*$
   values for the same target (Corollary 1).

4. **Practical constraints**: The scaling factor must lie within bounds that
   produce realistic item parameters.

5. **Root and item scope**: A finite curve may have several roots, and a
   fixed-form coefficient is not the same estimand as an average across random
   item forms. Both must be specified before a scale can be interpreted.

### The APC Approximation

The Analytic Pre-Calibration (APC) formula provides a rough initial estimate
under Gaussian Rasch assumptions. With $\theta \sim N(0,1)$ and
$\beta \sim N(0, \sigma^2_\beta)$, the expected item information involves
the logistic-normal convolution constant:

$$
\kappa(\sigma^2) = \int_{-\infty}^{\infty}
\frac{e^z}{(1+e^z)^2}\, \phi(z; 0, \sigma^2)\, dz
\approx \frac{0.25}{\sqrt{1 + \sigma^2 \pi^2 / 3}}
\qquad \text{(24)}
$$

where $\sigma^2 = 1 + \sigma^2_\beta$. The APC initial value is:

$$
c_{\text{APC}} = \sqrt{\frac{\rho^*}{I \cdot \kappa \cdot (1 - \rho^*)}}
\qquad \text{(25)}
$$

This approximation is typically within a factor of 2 of the true $c^*$ and
serves as a starting point for the stochastic approximation algorithm (SAC).


## Complete Notation Table

The following table summarizes all mathematical notation used in the package
and its documentation.

| Symbol | Description | Domain |
|:------:|:-----------|:------:|
| $\theta$ | Latent ability parameter | $\mathbb{R}$ |
| $G$ | Latent trait distribution | -- |
| $\sigma^2_\theta$ | Variance of $\theta$ under $G$ | $(0, \infty)$ |
| $\beta_i$ | Difficulty parameter for item $i$ | $\mathbb{R}$ |
| $\lambda_i$ | Discrimination parameter for item $i$ | $(0, \infty)$ |
| $\lambda_i^{(0)}$ | Baseline (unscaled) discrimination | $(0, \infty)$ |
| $g_i$ | 3PL lower asymptote (`guessing`) | $[0, 1)$ |
| $c$ | Global discrimination scaling factor | $(0, \infty)$ |
| $c^*$ | Calibrated scaling factor | $(0, \infty)$ |
| $I$ | Number of items | $\mathbb{N}$ |
| $M$ | Quadrature (Monte Carlo) sample size | $\mathbb{N}$ |
| $\psi(z)$ | Logistic function $1/(1+e^{-z})$ | $(0, 1)$ |
| $p_i(\theta)$ | $P(X_i = 1 \mid \theta)$; ICC | $(0, 1)$ |
| $\mathcal{J}_i(\theta)$ | Item information function | $(0, \infty)$ |
| $\mathcal{J}(\theta)$ | Test information function | $(0, \infty)$ |
| $\bar{\mathcal{J}}(c)$ | Average test information $\mathbb{E}_G[\mathcal{J}(\theta;c)]$ | $(0, \infty)$ |
| $\tilde{\rho}(c)$ | Average-information reliability | $(0, 1)$ |
| $\bar{w}(c)$ | MSEM-based reliability | $(0, 1)$ |
| $\rho^*$ | Target reliability | $(0, 1)$ |
| $\hat{\rho}_M(c)$ | Empirical reliability (EQC) | $(0, 1)$ |
| $\hat{\rho}_n(c)$ | Noisy reliability estimate (SAC) | $(0, 1)$ |
| $a_n$ | Step size at iteration $n$ | $(0, \infty)$ |
| $a$ | Step size constant (SAC) | $(0, \infty)$ |
| $A$ | Stabilization constant (SAC) | $[0, \infty)$ |
| $\gamma$ | Step decay exponent (SAC) | $(1/2, 1]$ |
| $B$ | Burn-in period (SAC) | $\mathbb{N}_0$ |
| $N$ | Total number of SAC iterations | $\mathbb{N}$ |
| $\bar{c}_N$ | Polyak-Ruppert average | $(0, \infty)$ |
| $\kappa$ | Logistic-normal convolution constant (APC) | $(0, 1/4)$ |
| $H$ | Item parameter distribution | -- |
| $\text{MSEM}$ | $\mathbb{E}[1/\mathcal{J}(\theta;c)]$ | $(0, \infty)$ |


## Summary of Theoretical Results

| Result | Statement | Implication |
|:-------|:----------|:------------|
| **Theorem 1** | $\tilde{\rho}(c) \geq \bar{w}(c)$ (Jensen) | Info metric is at least as large on the same population/grid basis |
| **Corollary 1** | $c^*_{\bar{w}} \geq c^*_{\tilde{\rho}}$ on a monotone MSEM branch | MSEM metric requires a weakly larger scale when the MSEM root is unique on the selected interval |
| **Proposition 1** | $\tilde{\rho}(c)$ is strictly increasing on regular calibration intervals | Existence and uniqueness for bracketed info targets |
| **Corollary 2** | Unique $c^*$ exists for $\rho^* \in (\rho_L,\rho_U)$ on such an interval | EQC root-finding is well-defined after feasibility checks |
| **Proposition 2** | $\bar{w}(c)$ is not always monotone | `c_bounds` critical for MSEM metric |
| **Proposition 3** | $\max \mathcal{J}_i = \lambda_i^2/4$ for 2PL | 2PL-only analytic upper bound |
| **Theorem 2** | EQC consistent for an isolated selected branch ($\hat{c}^*_M \to c^*$ a.s.) | Precision increases with $M$ under the stated conditions |
| **Theorem 3** | EQC asymptotically normal ($\sqrt{M}$ rate) | Quantifiable uncertainty |
| **Theorem 4** | SAC converges locally on an admissible selected branch, with PR averaging | Preflight and branch diagnostics are required |


## References

Lee, J.-H. (2026). Reliability-Targeted Simulation of Item Response Data:
Solving the Inverse Design Problem. arXiv:2512.16012v2.
https://doi.org/10.48550/arXiv.2512.16012

Robbins, H., & Monro, S. (1951). A stochastic approximation method. *The Annals
of Mathematical Statistics, 22*(3), 400--407.

Polyak, B. T., & Juditsky, A. B. (1992). Acceleration of stochastic
approximation by averaging. *SIAM Journal on Control and Optimization, 30*(4),
838--855.

Kushner, H. J., & Yin, G. G. (2003). *Stochastic Approximation and Recursive
Algorithms and Applications* (2nd ed.). Springer.

Brent, R. P. (1973). *Algorithms for Minimization Without Derivatives*.
Prentice-Hall.

Lord, F. M. (1980). *Applications of Item Response Theory to Practical Testing
Problems*. Erlbaum.

van der Linden, W. J. (Ed.). (2016). *Handbook of Item Response Theory* (Vols.
1--3). CRC Press.

Birnbaum, A. (1968). Some latent trait models and their use in inferring an
examinee's ability. In F. M. Lord & M. R. Novick (Eds.), *Statistical Theories
of Mental Test Scores* (pp. 395--479). Addison-Wesley.
